### The WHO dataset: Load Libraries and Data

Before starting, we need to load the required libraries and load the WHO dataset:

```{r}
#| output: false
library("tidyverse")
library("here")
library("cowplot")
library("patchwork")
library("krulRutils")
library("ISLR2")
library("magrittr")

options(scipen = 999)
data(who)
```

### Cleaning the WHO Dataset

We now proceed to clean the WHO dataset. The main problem is that it contains cryptic variables like `"new_sp_m014"`.

```{r}
who_tidy <- who %>%
  pivot_longer(
    cols = starts_with("new"),
    names_to = "key",
    values_to = "cases",
    values_drop_na = TRUE
  ) %>%
  mutate(
    key = if_else(
      startsWith(key, "newrel"),
      sub("newrel", "new_rel", key),
      key
    ),
    cases = as.integer(cases)
  ) %>%
  separate(key, into = c("new", "type", "sexage"), sep = "_") %>%
  separate(sexage, into = c("sex", "age"), sep = 1) %>%
  select(-new, -iso2, - iso3)  %T>%
  saveRDS(here("data", "who_tidy.rds")) %>%
  write_csv(here("data", "who_tidy.csv")) %>%
  glimpse()
```

### Lookup Tables for Factors

We now define lookup tables for `type`, `sex`, and `age`:

```{r}
who_type_lookup_tbl <- tibble(
  code = c("ep", "rel", "sn", "sp"),
  label = c(
    "Extrapulmonary TB",
    "Relapse case",
    "Smear-Negative pulmonary TB",
    "Smear-Positive pulmonary TB"
  )
)

who_sex_lookup_tbl <- tibble(
  code = c("f", "m"),
  label = c("Female", "Male")
)

who_age_lookup_tbl <- tibble(
  code = c("014", "1524", "2534", "3544", "4554", "5564", "65"),
  label = c("0-14", "15-24", "25-34", "35-44", "45-54", "55-64", "65+")
)
```

### Convert WHO Columns to Factors

We now convert the codes to factors using the lookup tables:

```{r}
who_factor <- who_tidy %>%
  convert_codes_to_factor(type, who_type_lookup_tbl, code, label) %>%
  convert_codes_to_factor(sex, who_sex_lookup_tbl, code, label) %>%
  convert_codes_to_factor(age, who_age_lookup_tbl, code, label) %>%
  glimpse()
```

### Bar Plot: Tuberculosis Cases by Type and Sex

First, we summarize the data:

```{r}
who_type_sex_tbl <- who_factor %>%
  count(type_factor, sex_factor, wt = cases, name = "cases")
```

Then, we create the plot:

```{r}
who_type_sex_plot <- who_type_sex_tbl %>%
  ggplot(aes(x = type_factor, y = cases, fill = sex_factor)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Female" = "#e7298a", "Male" = "#1f77b4")) +
  labs(
    title = "Tuberculosis Cases by Type and Sex",
    x = "Tuberculosis Type",
    y = "Cases",
    fill = "Sex"
  ) +
  theme_krul()
```

### Save and Display Plot

```{r}
#| fig-width: 12
#| fig-height: 6
#| fig-cap: "\\small TB cases are consistently higher among males across all types."
ggsave(
  filename = here("images", "who_plot.png"),
  plot = who_type_sex_plot,
  width = 12,
  height = 6,
  dpi = 300
)

who_type_sex_plot
```

### The Billboard Dataset: Tidying

We now clean the Billboard dataset, moving week columns into one.

```{r}
data(billboard)

billboard_tidy <- billboard %>%
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "ranking",
    values_drop_na = TRUE
  ) %>%
  mutate(
    week = if_else(startsWith(week, "wk"), sub("wk", "", week), week),
    week = as.integer(week),
    date = date.entered + weeks(week - 1)
  ) %>%
  arrange(date, ranking) %T>%
  saveRDS(here("data", "billboard_tidy.rds")) %>%
  write_csv(here("data", "billboard_tidy.csv"))
```

### Filter "Who Let The Dogs Out"

We filter and select the relevant data:

```{r}
dogs_out_tbl <- billboard_tidy %>%
  filter(track == "Who Let The Dogs Out") %>%
  arrange(week) %>%
  select(date, ranking)
```

### Plot Chart Performance

We now generate the plot:

```{r}
dogs_out_plot <- dogs_out_tbl %>%
  ggplot(aes(x = date, y = ranking)) +
  geom_line(color = "#1f77b4", linewidth = 1) +
  geom_point(color = "#d62728", size = 2) +
  scale_x_date(date_labels = "%B %Y") +
  scale_y_reverse(breaks = seq(0, 100, 10)) +
  labs(
    title = "Chart Performance of 'Who Let The Dogs Out'",
    x = "Date",
    y = "Billboard Hot 100 Ranking"
  ) +
  theme_krul()
```

### Save and Display the Chart

```{r}
#| fig-width: 12
#| fig-height: 6
#| fig-cap: "\\small The song peaked at rank 40 in the week of October 21, 2000, and stayed in the chart for several weeks."
ggsave(
  filename = here("images", "dogs_out_plot.png"),
  plot = dogs_out_plot,
  width = 12,
  height = 6,
  dpi = 300
)

dogs_out_plot
```
