# Skewness, Kurtosis and QQ Plots 

### Skewness, Kurtosis and QQ Plots

As it is well known, any normal distribution is completely determined by its mean and standard deviation.

\bigpause

However, many real-world datasets do not follow a normal distribution.

\bigpause

In this presentation, we will explore some other distributions and show how the concepts of skewness and kurtosis can help us understand its shape.

\bigpause

We will also introduce QQ plots, which are useful graphical tools for comparing the distribution of a dataset against a theoretical distribution, which is usually taken to be normal.

```{r}
#| output: false
#| echo: false
library("tidyverse")
library("here")
library("patchwork")
library("krulRutils")
library("ISLR2")
library("magrittr")
library("e1071") 

options(scipen = 999)
```

## Skewness 

### Skewness

\begin{definition}
The \alert{skewness} of a random variable $X$ is defined as the third standardized moment, given by
$$
\gamma_1 = \frac{\mu_3}{\sigma^3} = \frac{\mathbb{E}[(X - \mu)^3]}{\sigma^3},
$$
where $\mu = \mathbb{E}[X]$ is the mean and $\sigma = \sqrt{\mathbb{E}[(X - \mu)^2]}$ is the standard deviation of $X$.
\end{definition}

\pause

Skewness measures the asymmetry of a probability distribution.

\bigpause

A distribution is said to be \alert{positively skewed} (or right-skewed) if it has a longer tail on the right side, and \alert{negatively skewed} (or left-skewed) if it has a longer tail on the left side.

### Gamma Distribution

\begin{definition}
We say that a continuous random variable $X$ follows a \alert{Gamma distribution} with shape parameter $\alpha > 0$ and rate parameter $\beta > 0$, if its associated probability density function (PDF) is given by

\begin{equation} \label{eq:gamma_pdf}
f_X(x; \alpha, \beta) = 
\begin{cases}
\displaystyle \frac{\beta^\alpha}{\Gamma(\alpha)} \, x^{\alpha - 1} e^{-\beta x}, & x > 0, \\
0, & x \leq 0,
\end{cases}
\end{equation}

where $\Gamma(\alpha)$ denotes the Gamma function, defined as

$$
\Gamma(\alpha) = \int_0^\infty t^{\alpha - 1} e^{-t} \, dt.
$$

The parameters $\alpha$ and $\beta$ control the shape and rate of the distribution, respectively.
\end{definition}

###

Among other applications, the Gamma distribution is frequently used to model the number of total insurance claims over a given time period.

\pause

\begin{remark}
The skewness of a Gamma distribution is given by
$$
\gamma_1 = \frac{2}{\sqrt{\alpha}}.
$$
This means that the skewness is always positive, and it approaches zero as $\alpha$ increases. In other words, the distribution becomes more symmetric as $\alpha$ becomes larger.
\end{remark}

###

```{r}
#| fig-cap: "PDF of the gamma distribution with mean and median highlighted."
#| fig-width: 12
#| fig-height: 5
#| echo: false
# Parameters
alpha <- 2
beta <- 1

# Create sequence of x values
x_data <- seq(0, 10, length.out = 200)

# Calculate density values
gamma_distribution_tbl <- tibble(
  x_data = x_data,
  density = dgamma(x_data, shape = alpha, rate = beta)
)

# Calculate mean and median
mean_val <- alpha / beta
median_val <- qgamma(0.5, shape = alpha, rate = beta)

# Plot
gamma_distribution_tbl %>%
ggplot(aes(x = x_data, y = density)) +
  geom_line(
    color = c_palette["C blue"],
    linewidth = 1
  ) +
  geom_vline(
    xintercept = mean_val,
    color = c_palette["C red"],
    linetype = "dashed",
    linewidth = 1
  ) +
  geom_vline(
    xintercept = median_val,
    color = c_palette["C green"],
    linetype = "dotted",
    linewidth = 1
  ) +
  annotate(
    "text",
    x = mean_val,
    y = 0.35,
    label = "Mean",
    color = c_palette["C red"],
    hjust = -0.1
  ) +
  annotate(
    "text",
    x = median_val,
    y = 0.35,
    label = "Median",
    color = c_palette["C green"],
    hjust = 1.1
  ) +
  labs(
    title = paste(
      "Gamma Distribution PDF (shape =", alpha, ", rate =", beta, ")"
    ),
    x = "x",
    y = "Density"
  ) +
  theme_krul()
```

## Kurtosis 

### Kurtosis

\begin{definition}
The \alert{kurtosis} of a random variable $X$ is defined as the fourth standardized moment, given by
$$
\beta_2 = \frac{\mu_4}{\sigma^4} = \frac{\mathbb{E}[(X - \mu)^4]}{\sigma^4},
$$
where $\mu = \mathbb{E}[X]$ is the mean and $\sigma = \sqrt{\mathbb{E}[(X - \mu)^2]}$ is the standard deviation of $X$.
\end{definition}

\pause

Kurtosis measures the "tailedness" of a probability distribution.

\pause

A distribution with high kurtosis has heavier tails and a sharper peak than a normal distribution, while a distribution with low kurtosis has lighter tails and a flatter peak.

\pause

The kurtosis of a normal distribution is 3, which is why frequently we are interested in computing the \emph{excess kurtosis} which is given by:
$$
\gamma_2 = \beta_2 - 3.
$$

### Laplace Distribution

\begin{definition}
We say that a continuous random variable $X$ follows a \alert{Laplace distribution} with location parameter $\mu \in \mathbb{R}$ and scale parameter $b > 0$, if its associated probability density function (PDF) is given by
\begin{equation} \label{eq:laplace_pdf}
f_X(x; \mu, b) = \frac{1}{2b} \exp\left(-\frac{|x - \mu|}{b}\right), \quad x \in \mathbb{R}.
\end{equation}
The parameters $\mu$ and $b$ control the location and scale of the distribution, respectively.
\end{definition}

###

Among other applications, the Laplace distribution is often used in image processing and computer vision, particularly in modeling noise in images.

\pause

\begin{remark} The excess kurtosis of a Laplace distribution is given by
$$
\gamma_2 = 3.
$$
This means that the Laplace distribution has heavier tails than a normal distribution.
\end{remark}

###

```{r}
#| fig-cap: "PDF of the Laplace distribution with mean zero and variance one."
#| fig-width: 12
#| fig-height: 5
#| echo: false

# Parameters
mu <- 0      # mean (location)
b <- 1/sqrt(2)       # scale

# Sequence of x values covering enough range to see tails
x_data <- seq(-5, 5, length.out = 300)

# Laplace PDF function
dlaplace <- function(x, mu, b) {
  return(1/(2*b) * exp(-abs(x - mu)/b))
}

# Create tibble with PDF values
laplace_distribution_tbl<- tibble(
  x_data = x_data,
  density = dlaplace(x_data, mu, b)
)

# Plot PDF
laplace_distribution_tbl %>% 
  ggplot(aes(x = x_data, y = density)) +
  geom_line(color = c_palette["C blue"], linewidth = 1) +
  labs(
    title = "Laplace Distribution PDF (mean = 0, var = 1)",
    x = "x",
    y = "Density"
  ) +
  theme_krul()
```

## QQ Plots

### QQ Plots

QQ plots, or quantile-quantile plots, are graphical tools used to compare the distribution of a dataset against a theoretical distribution, such as the normal distribution.

\bigpause

They are particularly useful for assessing the normality of the data.

### Constructing a QQ Plot

To construct a QQ plot, you should follow these steps:
\begin{enumerate}
    \item  \alert{Sort the sample data:}  
    Arrange the data in increasing order:  
    \[
    x_{(1)} \leq x_{(2)} \leq \cdots \leq x_{(n)}.
    \]
    
    \item  \alert{Calculate plotting positions (empirical quantiles):}  
    For each ordered value \( x_{(i)} \), compute the corresponding probability  
    \[
    p_i = \frac{2i - 1}{2n}, \quad i = 1, 2, \ldots, n.
    \]
    
\seti
 \end{enumerate}       

### Constructing a QQ Plot

\begin{enumerate}

\conti

    \item  \alert{Find theoretical quantiles of the normal distribution:}  
    Compute the theoretical quantiles from the inverse cumulative distribution function (CDF) of the normal distribution:  
    \[
    q_i = \Phi^{-1}(p_i),
    \]
    where \( \Phi^{-1} \) is the quantile function of the standard normal distribution.  



    \item  \alert{Plot the points:}  
    Plot the pairs \( (q_i, x_{(i)}) \) with \( q_i \) on the x-axis and \( x_{(i)} \) on the y-axis.
    
    \item  \alert{Add a reference line:}  
    Typically, we add a line through the first and third quartiles to help assess normality. 

\end{enumerate}

The closer the points lie to this line, the more the sample resembles the specified normal distribution.


###
```{r}
#| output: false
#| echo: false
set.seed(123)  # for reproducibility

# Sample size 
n <- 300 
# Generate random samples from the Gamma distribution 
gamma_sample <- rgamma(n, shape = alpha, rate = beta) 

gamma_sample_tbl <- tibble(
  sample = gamma_sample
)

# Generate uniform random numbers in (0,1)
u <- runif(n)

# Apply inverse CDF of Laplace
laplace_sample <- ifelse(u < 0.5,
                         mu + b * log(2*u),
                         mu - b * log(2*(1 - u)))

laplace_sample_tbl <- tibble(
  sample = laplace_sample
)
```

```{r}
#| fig-cap: "QQ plot and histogram of the Gamma distribution samples." 
#| fig-width: 12 
#| fig-height: 5
#| echo: false
gamma_qq_plot <- gamma_sample_tbl %>%
  ggplot(aes(sample = sample)) +
  geom_qq(
    color = c_palette["C red"],
    size = 0.3
  ) +
  geom_qq_line(
    color = c_palette["C blue"],
    linewidth = 0.5
  ) +
  labs(
    title = "QQ Plot",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_krul()

gamma_histogram <- gamma_sample_tbl %>%
  ggplot(aes(x = sample)) +
  geom_histogram(
    bins = 30,
    fill = c_palette["C blue"],
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Histogram",
    x = "Sample Values",
    y = "Density"
  ) +
  theme_krul()

gamma_plot <- gamma_qq_plot + gamma_histogram +
  plot_layout(ncol = 2) +

  plot_annotation(
    title = "Gamma Distribution: QQ Plot and Histogram",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )

gamma_plot
```


###
```{r}
#| fig-cap: "QQ plot and histogram of the Laplace distribution samples."
#| fig-width: 12 
#| fig-height: 5 
#| echo: false

laplace_qq_plot <- laplace_sample_tbl %>%
  ggplot(aes(sample = sample)) +
  geom_qq(
    color = c_palette["C red"],
    size = 0.3
  ) +
  geom_qq_line(
    color = c_palette["C blue"],
    linewidth = 0.5
  ) +
  labs(
    title = "QQ Plot",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_krul()

laplace_histogram <- laplace_sample_tbl %>%
  ggplot(aes(x = sample)) +
  geom_histogram(
    bins = 30,
    fill = c_palette["C blue"],
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Histogram",
    x = "Sample Values",
    y = "Density"
  ) +
  theme_krul()
laplace_plot <- laplace_qq_plot + laplace_histogram +
  plot_layout(ncol = 2) +
  
  plot_annotation(
    title = "Laplace Distribution: QQ Plot and Histogram",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )
laplace_plot 
```

## Sample Skewness and Kurtosis 

### Sample Skewness and Kurtosis

So far we have defined skewness and kurtosis for random variables.

\bigpause

However, in practice, we often work with samples rather than entire populations.

\bigpause

Therefore, we need to define sample skewness and sample kurtosis.

###

\begin{definition}
Let $x_1, x_2, \dots, x_n$ be a sample of size $n$. 
The \emph{sample mean} $\bar{x}$ and \emph{sample variance} $s^{2}$ are given by
\[
\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i,
\qquad
s^{2} = \frac{1}{n-1} \sum_{i=1}^n (x_i - \bar{x})^2.
\]
The \emph{sample skewness} $g_1$ given by
\[
g_1 = \frac{n}{(n-1)(n-2)} \sum_{i=1}^n \left( \frac{x_i - \bar{x}}{s} \right)^3.
\]
The \emph{sample excess kurtosis} $g_2$ 
is given by
$$
g_2 = \frac{n(n+1)}{(n-1)(n-2)(n-3)}
\sum_{i=1}^n \left( \frac{x_i - \bar{x}}{s} \right)^4
- \frac{3(n-1)^2}{(n-2)(n-3)}.
$$
\end{definition}

### 


As an example, we will compute the sample skewness and sample excess kurtosis of the Gamma and Laplace samples we showed earlier.
\bigpause 

For the Gamma sample we have:
```{r}
#| echo: false
# Compute sample skewness and excess kurtosis for Gamma distribution 
gamma_sample_tbl %>%
  summarise(
    sample_skewness = skewness(sample, type = 2),
    sample_excess_kurtosis = kurtosis(sample, type = 2) - 3
  ) %>%
glimpse()
```


\bigpause


For the Laplace sample we have:
```{r}
#| echo: false
# Compute sample skewness and excess kurtosis for Laplace distribution 
laplace_sample_tbl %>%
  summarise(
    sample_skewness = skewness(sample, type = 2),
    sample_excess_kurtosis = kurtosis(sample, type = 2)
  ) %>%
glimpse()
```


## Activity: Assessing Normality

### Activity: Assessing Normality

For each of the following distributions, construct a sample of size 300 and compute their associated sample Skewness and Excess Kurtosis. After that, construct and compare their associated QQ plots and histograms.

\begin{enumerate}
    \item Exponential distribution with rate parameter $\lambda = 1$.
    \item Uniform distribution on the interval $[0, 1]$.
    \item Beta distribution with shape parameters $\alpha = 2$ and $\beta = 5$.
    \item Cauchy distribution with location parameter $x_0 = 0$ and scale parameter $\gamma = 1$.
    \item Chi-squared distribution with degrees of freedom $k = 3$.
\end{enumerate}
