# Asimetría, curtosis y gráficos QQ 

### Asimetría, curtosis y gráficos QQ

Como es bien sabido, cualquier distribución normal queda completamente determinada por su media y su desviación estándar.

\bigpause

Sin embargo, muchos conjuntos de datos del mundo real no siguen una distribución normal.

\bigpause

En esta presentación exploraremos otras distribuciones y mostraremos cómo los conceptos de asimetría y curtosis ayudan a entender su forma.

\bigpause

También introduciremos los gráficos QQ, herramientas útiles para comparar la distribución de un conjunto de datos contra una distribución teórica (usualmente la normal).

```{r}
#| output: false
#| echo: false
library(tidyverse)
library(broom)
library(car)
library(here)
library(patchwork)
library(krulRutils)
library(ISLR2)
library(magrittr)
library(e1071)

options(scipen = 999)
```

## Asimetría 

### Asimetría

\begin{definition}
La \alert{asimetría} de una variable aleatoria $X$ es el tercer momento estandarizado, dado por
$$
\gamma_1 = \frac{\mu_3}{\sigma^3} = \frac{\mathbb{E}[(X - \mu)^3]}{\sigma^3},
$$
donde $\mu = \mathbb{E}[X]$ es la media y $\sigma = \sqrt{\mathbb{E}[(X - \mu)^2]}$ es la desviación estándar de $X$.
\end{definition}

\pause

La asimetría mide la falta de simetría de una distribución de probabilidad.

\bigpause

Una distribución es \alert{positivamente asimétrica} (o con cola a la derecha) si tiene una cola más larga a la derecha, y \alert{negativamente asimétrica} (o con cola a la izquierda) si la cola más larga está a la izquierda.

### Distribución Gamma

\begin{definition}
Decimos que una variable aleatoria continua $X$ sigue una \alert{distribución Gamma} con parámetro de forma $\alpha > 0$ y de tasa $\beta > 0$, si su función de densidad (PDF) es

\begin{equation} \label{eq:gamma_pdf}
f_X(x; \alpha, \beta) = 
\begin{cases}
\displaystyle \frac{\beta^\alpha}{\Gamma(\alpha)} \, x^{\alpha - 1} e^{-\beta x}, & x > 0, \\
0, & x \leq 0,
\end{cases}
\end{equation}

donde $\Gamma(\alpha)$ denota la función Gamma, definida como

$$
\Gamma(\alpha) = \int_0^\infty t^{\alpha - 1} e^{-t} \, dt.
$$

Los parámetros $\alpha$ y $\beta$ controlan la forma y la tasa de la distribución, respectivamente.
\end{definition}

###

Entre otras aplicaciones, la distribución Gamma se usa con frecuencia para modelar el número total de reclamaciones de seguros en un periodo dado.

\pause

\begin{remark}
The skewness of a Gamma distribution is given by
$$
\gamma_1 = \frac{2}{\sqrt{\alpha}}.
$$
This means that the skewness is always positive, and it approaches zero as $\alpha$ increases. In other words, the distribution becomes more symmetric as $\alpha$ becomes larger.
\end{remark}

###

```{r}
#| fig-cap: "PDF de la distribución Gamma con media y mediana destacadas."
#| fig-width: 12
#| fig-height: 5
#| echo: false
# Parameters
alpha <- 2
beta <- 1

# Create sequence of x values
x_data <- seq(0, 10, length.out = 200)

# Calculate density values
gamma_distribution_tbl <- tibble(
  x_data = x_data,
  density = dgamma(x_data, shape = alpha, rate = beta)
)

# Calculate mean and median
gamma_mean <- alpha / beta
gamma_median <- qgamma(0.5, shape = alpha, rate = beta)
gamma_sd <- sqrt(alpha) / beta

# Plot
gamma_distribution_plot <- gamma_distribution_tbl %>%
  ggplot(aes(x = x_data, y = density)) +
  geom_line(
    color = c_palette["C blue"],
    linewidth = 1
  ) +
  geom_vline(
    xintercept = gamma_mean,
    color = c_palette["C red"],
    linetype = "dashed",
    linewidth = 1
  ) +
  geom_vline(
    xintercept = gamma_median,
    color = c_palette["C green"],
    linetype = "dotted",
    linewidth = 1
  ) +
  annotate(
    "text",
    x = gamma_mean,
    y = 0.35,
    label = "Media",
    color = c_palette["C red"],
    hjust = -0.1
  ) +
  annotate(
    "text",
    x = gamma_median,
    y = 0.35,
    label = "Mediana",
    color = c_palette["C green"],
    hjust = 1.1
  ) +
  labs(
    title = paste(
      "PDF de la distribución Gamma (shape =", alpha, ", rate =", beta, ")"
    ),
    x = "x",
    y = "Densidad"
  ) +
  theme_krul()

gamma_distribution_plot
```

## Curtosis 

### Curtosis

\begin{definition}
La \alert{curtosis} de una variable aleatoria $X$ es el cuarto momento estandarizado, dado por
$$
\beta_2 = \frac{\mu_4}{\sigma^4} = \frac{\mathbb{E}[(X - \mu)^4]}{\sigma^4},
$$
donde $\mu = \mathbb{E}[X]$ es la media y $\sigma = \sqrt{\mathbb{E}[(X - \mu)^2]}$ es la desviación estándar de $X$.
\end{definition}

\pause

La curtosis mide el grosor de las colas de una distribución.

\pause

Una distribución con alta curtosis tiene colas más pesadas y un pico más agudo que una normal; con baja curtosis, colas más ligeras y un pico más plano.

\pause

La curtosis de una normal es 3; por ello suele calcularse la \alert{curtosis en exceso} dada por:
$$
\gamma_2 = \beta_2 - 3.
$$

### Distribución de Laplace

\begin{definition}
Decimos que una variable aleatoria continua $X$ sigue una \alert{distribución de Laplace} con parámetro de localización $\mu \in \mathbb{R}$ y escala $b > 0$, si su PDF es
\begin{equation} \label{eq:laplace_pdf}
f_X(x; \mu, b) = \frac{1}{2b} \exp\left(-\frac{|x - \mu|}{b}\right), \quad x \in \mathbb{R}.
\end{equation}
Los parámetros $\mu$ y $b$ controlan la localización y la escala de la distribución, respectivamente.
\end{definition}

###

Entre otras aplicaciones, la distribución de Laplace se utiliza en procesamiento de imágenes para modelar ruido.

\pause

\begin{remark} La curtosis en exceso de una distribución de Laplace es
$$
\gamma_2 = 3.
$$
Esto significa que la distribución de Laplace tiene colas más pesadas que la normal.
\end{remark}

###

```{r}
#| fig-cap: "PDF de la distribución de Laplace con media cero y varianza uno."
#| fig-width: 12
#| fig-height: 5
#| echo: false

# Parameters
mu <- 0 # mean (location)
b <- 1 / sqrt(2) # scale

# Sequence of x values covering enough range to see tails
x_data <- seq(-5, 5, length.out = 300)

# Laplace PDF function
dlaplace <- function(x, mu, b) {
  return(1 / (2 * b) * exp(-abs(x - mu) / b))
}

# Create tibble with PDF values
laplace_distribution_tbl <- tibble(
  x_data = x_data,
  density = dlaplace(x_data, mu, b)
)

# Plot PDF
laplace_distribution_tbl %>%
  ggplot(aes(x = x_data, y = density)) +
  geom_line(color = c_palette["C blue"], linewidth = 1) +
  labs(
    title = "Laplace Distribution PDF (mean = 0, var = 1)",
    x = "x",
    y = "Density"
  ) +
  theme_krul()
```

## QQ Plots

### QQ Plots

QQ plots, or quantile-quantile plots, are graphical tools used to compare the distribution of a dataset against a theoretical distribution, such as the normal distribution.

\bigpause

They are particularly useful for assessing the normality of the data.

### Constructing a QQ Plot

To construct a QQ plot, you should follow these steps:
\begin{enumerate}
    \item  \alert{Sort the sample data:}  
    Arrange the data in increasing order:  
    \[
    x_{(1)} \leq x_{(2)} \leq \cdots \leq x_{(n)}.
    \]
    
    \item  \alert{Calculate plotting positions (empirical quantiles):}  
    For each ordered value \( x_{(i)} \), compute the corresponding probability  
    \[
    p_i = \frac{2i - 1}{2n}, \quad i = 1, 2, \ldots, n.
    \]
    
\seti
 \end{enumerate}       

### Constructing a QQ Plot

\begin{enumerate}

\conti

    \item  \alert{Find theoretical quantiles of the normal distribution:}  
    Compute the theoretical quantiles from the inverse cumulative distribution function (CDF) of the normal distribution:  
    \[
    q_i = \Phi^{-1}(p_i),
    \]
    where \( \Phi^{-1} \) is the quantile function of the standard normal distribution.  



    \item  \alert{Plot the points:}  
    Plot the pairs \( (q_i, x_{(i)}) \) with \( q_i \) on the x-axis and \( x_{(i)} \) on the y-axis.
    
    \item  \alert{Add a reference line:}  
    Typically, we add a line through the first and third quartiles to help assess normality. 

\end{enumerate}

The closer the points lie to this line, the more the sample resembles the specified normal distribution.


###
```{r}
#| output: false
#| echo: false
set.seed(123) # for reproducibility

# Sample size
n <- 300
# Generate random samples from the Gamma distribution
gamma_sample <- rgamma(n, shape = alpha, rate = beta)

gamma_sample_tbl <- tibble(
  sample = gamma_sample
)

# Generate uniform random numbers in (0,1)
u <- runif(n)

# Apply inverse CDF of Laplace
laplace_sample <- ifelse(u < 0.5,
  mu + b * log(2 * u),
  mu - b * log(2 * (1 - u))
)

laplace_sample_tbl <- tibble(
  sample = laplace_sample
)
```

```{r}
#| fig-cap: "QQ plot and histogram of the Gamma distribution samples."
#| fig-width: 12
#| fig-height: 5
#| echo: false
gamma_qq_plot <- gamma_sample_tbl %>%
  ggplot(aes(sample = sample)) +
  geom_qq(
    color = c_palette["C red"],
    size = 0.3
  ) +
  geom_qq_line(
    color = c_palette["C blue"],
    linewidth = 0.5
  ) +
  labs(
    title = "QQ Plot",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_krul()

gamma_histogram <- gamma_sample_tbl %>%
  ggplot(aes(x = sample)) +
  geom_histogram(
    bins = 30,
    fill = c_palette["C blue"],
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Histogram",
    x = "Sample Values",
    y = "Density"
  ) +
  theme_krul()

gamma_plot <- gamma_qq_plot + gamma_histogram +
  plot_layout(ncol = 2) +

  plot_annotation(
    title = "Gamma Distribution: QQ Plot and Histogram",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )

gamma_plot
```


###
```{r}
#| fig-cap: "QQ plot and histogram of the Laplace distribution samples."
#| fig-width: 12
#| fig-height: 5
#| echo: false

laplace_qq_plot <- laplace_sample_tbl %>%
  ggplot(aes(sample = sample)) +
  geom_qq(
    color = c_palette["C red"],
    size = 0.3
  ) +
  geom_qq_line(
    color = c_palette["C blue"],
    linewidth = 0.5
  ) +
  labs(
    title = "QQ Plot",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_krul()

laplace_histogram <- laplace_sample_tbl %>%
  ggplot(aes(x = sample)) +
  geom_histogram(
    bins = 30,
    fill = c_palette["C blue"],
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Histogram",
    x = "Sample Values",
    y = "Density"
  ) +
  theme_krul()
laplace_plot <- laplace_qq_plot + laplace_histogram +
  plot_layout(ncol = 2) +

  plot_annotation(
    title = "Laplace Distribution: QQ Plot and Histogram",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )
laplace_plot
```

## Sample Skewness and Kurtosis 

### Sample Skewness and Kurtosis

So far we have defined skewness and kurtosis for random variables.

\bigpause

However, in practice, we often work with samples rather than entire populations.

\bigpause

Therefore, we need to define sample skewness and sample kurtosis.

###

\begin{definition}
Let $x_1, x_2, \dots, x_n$ be a sample of size $n$. 
The \alert{sample mean} $\bar{x}$ and \alert{sample variance} $s^{2}$ are given by
\[
\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i,
\qquad
s^{2} = \frac{1}{n-1} \sum_{i=1}^n (x_i - \bar{x})^2.
\]
The \alert{sample skewness} $g_1$ given by
\[
g_1 = \frac{n}{(n-1)(n-2)} \sum_{i=1}^n \left( \frac{x_i - \bar{x}}{s} \right)^3.
\]
The \alert{sample excess kurtosis} $g_2$ 
is given by
$$
g_2 = \frac{n(n+1)}{(n-1)(n-2)(n-3)}
\sum_{i=1}^n \left( \frac{x_i - \bar{x}}{s} \right)^4
- \frac{3(n-1)^2}{(n-2)(n-3)}.
$$
\end{definition}

### 


As an example, we will compute the sample skewness and sample excess kurtosis of the Gamma and Laplace samples we showed earlier.
\bigpause 

For the Gamma sample we have:
```{r}
#| echo: false
# Compute sample skewness and excess kurtosis for Gamma distribution
gamma_sample_tbl %>%
  summarise(
    sample_skewness = skewness(sample, type = 2),
    sample_excess_kurtosis = kurtosis(sample, type = 2) - 3
  ) %>%
  glimpse()
```


\bigpause


For the Laplace sample we have:
```{r}
#| echo: false
# Compute sample skewness and excess kurtosis for Laplace distribution
laplace_sample_tbl %>%
  summarise(
    sample_skewness = skewness(sample, type = 2),
    sample_excess_kurtosis = kurtosis(sample, type = 2)
  ) %>%
  glimpse()
```


# Scaling and Standardization 

### Scaling and Standardization

In many situations, it could happen that the units we are using to measure different variables in our data are not compatible.

\bigpause 

For example, we could have a dataset that contains the height of individuals in centimeters and their weight in kilograms.

\bigpause 

In such cases, it is often useful to \alert{scale} the variables so that they are comparable. 

## Scaling Random Variables

### Scaling Random Variables

\begin{definition} 
An \alert{affine transformation} of a random variable $X$ is a transformation of the form 
$$
Y = aX + b
$$
where $a$ and $b$ are constants.
\end{definition}

\pause

\begin{remark}
In the context of statistics, affine transformations are often refered to as \alert{scaling} (even when they also include a translational component).
\end{remark}


###

Now assume that $X$ is a random variable with mean $\mu$ and standard deviation $\sigma$.

\bigpause 

Then the random variable
$$ 
Y = aX + b 
$$
will have mean $a\mu + b$ and standard deviation $|a|\sigma$.

\bigpause 

In particular, if we choose $a = 1/\sigma$ and $b = -\mu/\sigma$, then the random variable
$$
Z = \frac{X - \mu}{\sigma}
$$
will have mean $0$ and standard deviation $1$.

### 
In particular, if we choose $a = 1/\sigma$ and $b = -\mu/\sigma$, then the random variable
$$
Z = \frac{X - \mu}{\sigma}
$$
will have mean $0$ and standard deviation $1$.

\bigpause

This particular affine transformation is called a \alert{standardization}, and the resulting random variable $Z$ is called the \alert{standardized} version of $X$. $Z$ is also sometimes called the \alert{z-score} of $X$.

\pause

\begin{remark}
Notice that if $X$ is not normally distributed, then $Z$ will \alert{not} be the standard normal distribution. Standardization does not make a non-normal distribution normal.
\end{remark}

###

\begin{observation}
Notice that, by definition, 
$$
\gamma_{1}(Z) =\mathbb{E}[Z^{3}] = \frac{\mathbb{E}[(X - \mu)^3]}{\sigma^3} = \gamma_{1}(X)
$$
and
$$
\gamma_{2}(Z) =\mathbb{E}[Z^{4}] - 3 = \frac{\mathbb{E}[(X - \mu)^4]}{\sigma^4} -3 = \gamma_{2}(X).
$$
\end{observation}


### 

As an example, we will consider again the Gamma distrubution that we introduced earlier. In this case we have that 
$$
\mu = \frac{\alpha}{\beta} \quad \text{and} \quad \sigma = \frac{\sqrt{\alpha}}{\beta}. 
$$
\pause 
Using this information, we can generate the plot of standadized version of the Gamma distribution and compare it with the original one.

###
```{r}
#| fig-cap: "Comparison of the probability density function (PDF) of the original and standardized Gamma distribution."
#| fig-width: 12
#| fig-height: 5
#| echo: false

# Calculate density values
z_gamma_distribution_tbl <- gamma_distribution_tbl %>%
  mutate(
    z_data = (x_data - gamma_mean) / gamma_sd,
    z_density = dgamma(x_data, shape = alpha, rate = beta) * gamma_sd
  ) %>%
  select(z_data, z_density)

# Calculate mean and median
z_gamma_mean <- 0
z_gamma_median <- (gamma_median - gamma_mean) / gamma_sd

# Plot
z_gamma_distribution_plot <- z_gamma_distribution_tbl %>%
  ggplot(aes(x = z_data, y = z_density)) +
  geom_line(
    color = c_palette["C blue"],
    linewidth = 1
  ) +
  geom_vline(
    xintercept = z_gamma_mean,
    color = c_palette["C red"],
    linetype = "dashed",
    linewidth = 1
  ) +
  geom_vline(
    xintercept = z_gamma_median,
    color = c_palette["C green"],
    linetype = "dotted",
    linewidth = 1
  ) +
  annotate(
    "text",
    x = z_gamma_mean,
    y = 0.55,
    label = "Mean",
    color = c_palette["C red"],
    hjust = -0.1
  ) +
  annotate(
    "text",
    x = z_gamma_median,
    y = 0.55,
    label = "Median",
    color = c_palette["C green"],
    hjust = 1.1
  ) +
  coord_cartesian(ylim = c(0, 0.6)) +
  labs(
    title = paste(
      "Standardized"
    ),
    x = "X",
    y = "Density"
  ) +
  theme_krul()

gamma_distribution_plot <- gamma_distribution_plot +
  labs(
    title = "Original"
  )

combined_gamma_plot <- gamma_distribution_plot + z_gamma_distribution_plot +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Gamma Distribution: Original and Standardized",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )

combined_gamma_plot
```


## Scaling Samples

### Scaling Samples 

For a given a sample, it is not always known what the underlying distribution is.

\bigpause 

In particular, we may not know what the mean and the standard deviation of the underlying population are.

\bigpause 

However, we can still try to scale our data but we should keep in mind that in this case there is more than one scaling method.

###
The most common are:

\begin{enumerate}
    \item \alert{Standardization}: This method uses the sample mean and sample standard deviation to scale the data. The standardized sample is given by
    $$
    T = \frac{X - \bar{X}}{S},
    $$
    where $\bar{X}$ is the sample mean and $S$ is the sample standard deviation. 
    
    \item \alert{Min-Max Scaling}: This method scales the data to a fixed range, usually [0, 1]. The scaled sample is given by
    $$
    T = \frac{X - \min(X)}{\max(X) - \min(X)},
    $$
    where $\min(X)$ and $\max(X)$ are the minimum and maximum values of the sample, respectively. This method is very sensitive to outliers, as it uses the minimum and maximum values of the sample.
\seti 
\end{enumerate} 

###
\begin{enumerate}
\conti

    \item \alert{Robust Scaling}: This method scales the data using the median and interquartile range (IQR). The scaled sample is given by
    $$
    T = \frac{X - \text{median}(X)}{\text{IQR}(X)},
    $$
    where $\text{IQR}(X) = Q_3 - Q_1$ is the interquartile range of the sample. This method is more robust to outliers than min-max scaling, as it uses the median and IQR instead of the minimum and maximum values.

    \item \alert{Normalization}: This method simply divides each data point by the corresponding vector norm. The normalized sample is given by
    $$
    T = \frac{X}{\|X\|},
    $$
    where $\|X\|$ is the vector norm of the sample. It is not used as often as the previous methods.
\end{enumerate}

###

\begin{remark} 
As we mentioned earlier, scaling does not make a non-normal distribution normal. In particular, the sample skewness and sample excess kurtosis of the scaled sample will be the same as those of the original sample.
\end{remark}

\pause

As an example, we will consider the different scaling methods applied to the Gamma sample we generated earlier.  


###

```{r}
#| fig-cap: "Comparison of different scaling methods applied to the Gamma distribution sample."
#| fig-width: 14
#| fig-height: 7
#| echo: false

scale_lookup_tbl <- tibble(
  code = c(
    "sample",
    "sample_standard",
    "sample_min_max",
    "sample_robust",
    "sample_normal"
  ),
  label = c(
    "Original Sample",
    "Standardized",
    "Min-Max Scaled",
    "Robust Scaled",
    "Normalized"
  )
)

scaled_gamma_sample_tbl <- gamma_sample_tbl %>%
  mutate(
    sample_standard = standard_scale(sample),
    sample_min_max = min_max_scale(sample),
    sample_robust = robust_scale(sample),
    sample_normal = normal_scale(sample)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "scaling_method",
    values_to = "sample_scaled"
  ) %>%
  convert_codes_to_factor(
    code_col = scaling_method,
    lookup_tbl = scale_lookup_tbl,
    lookup_code_col = code,
    lookup_label_col = label,
    new_factor_col_name = scaling_method_factor
  )

scaled_gamma_qq_plot <- scaled_gamma_sample_tbl %>%
  ggplot(aes(sample = sample_scaled)) +
  facet_wrap(
    ~scaling_method_factor,
    nrow = 5,
    scales = "free"
  ) +
  geom_qq(
    color = c_palette["C red"],
    size = 0.3
  ) +
  geom_qq_line(
    color = c_palette["C blue"],
    linewidth = 0.5
  ) +
  labs(
    title = "QQ Plots",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_krul()

scaled_gamma_histogram <- scaled_gamma_sample_tbl %>%
  ggplot(aes(x = sample_scaled)) +
  facet_wrap(
    ~scaling_method_factor,
    nrow = 5,
    scales = "free"
  ) +
  geom_histogram(
    bins = 30,
    fill = c_pal("C blue"),
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Histograms",
    x = "Sample Values",
    y = "Density"
  ) +
  theme_krul()

scaled_gamma_plot <- scaled_gamma_qq_plot + scaled_gamma_histogram +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Comparison of Scaling Methods Applied to the Gamma Sample",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )

scaled_gamma_plot
```

# Activity: Visualizing the Non-Normality of Samples

### Activity: Visualizing the Non-Normality of Samples

For each of the following distributions, construct a sample of size 300 and compute their associated sample Skewness and Kurtosis. After that, construct and compare their associated QQ plots and histograms.

\begin{enumerate}
    \item Exponential distribution with rate parameter $\lambda = 1$.
    \item Uniform distribution on the interval $[0, 1]$.
    \item Beta distribution with shape parameters $\alpha = 2$ and $\beta = 5$.
    \item Cauchy distribution with location parameter $x_0 = 0$ and scale parameter $\gamma = 1$.
    \item Chi-squared distribution with degrees of freedom $k = 3$.
\end{enumerate}

# Normality Testing

### Normality Testing

Although QQ plots are a good way to visually assess the normality of a dataset, they can be subjective and imprecise.

\bigpause 

Therefore, it becomes necessary to develop more objective and precise statistical tests to determine whether the data we are analyzing follows a normal distribution.

\bigpause 

We will now explore some commonly used statistical tests for normality and discuss some transformations that can be performed on our data to make it more normal-like. 

## Normality Tests 

### Normality Tests

There are several statistical tests that can be used to determine whether a dataset follows a normal distribution. Some of the most commonly used tests include:

\begin{enumerate}
\item \alert{Shapiro-Wilk test}: This test is based on the correlation between the data and the corresponding normal scores. It is a powerful test for small sample sizes (less than 50) and is widely used in practice. However, it can be overly sensitive to small deviations from normality in larger sample sizes. 

\item \alert{Kolmogorov-Smirnov test}: This test compares the empirical distribution function of the data with the cumulative distribution function of the normal distribution. It is a non-parametric test and can be used for any sample size. However, it is less powerful than the Shapiro-Wilk test for small sample sizes. It is also sensitive to scale and location parameters, which can affect the results of the test.
\seti 
\end{enumerate} 


###

\begin{enumerate}

\conti

\item \alert{Anderson-Darling test}: This test is similar to the Kolmogorov-Smirnov test but gives more weight to the tails of the distribution. It is a powerful test for small sample sizes and is widely used in practice.

\item \alert{Jarque-Bera test}: This test is based on the skewness and kurtosis of the data. It is a powerful test for large sample sizes and is widely used in practice. However, it can be overly sensitive to small deviations from normality in smaller sample sizes.
\end{enumerate}

We will now apply these tests to the gamma sample data we created in the previous chapter:

###

```{r}
#| output: false
#| echo: false
normality_tests_lookup_tbl <- tibble(
  code = c("shapiro", "ks", "ad", "jb"),
  label = c(
    "Shapiro-Wilk",
    "Kolmogorov-Smirnov",
    "Anderson-Darling",
    "Jarque-Bera"
  ),
)

normality_test_gamma_sample_tbl <- gamma_sample_tbl %>%
  summarise(
    shapiro = tidy_shapiro_test(data = ., value_col = sample)$p_value,
    ks = tidy_ks_test(data = ., value_col = sample)$p_value,
    ad = tidy_ad_test(data = ., value_col = sample)$p_value,
    jb = tidy_jb_test(data = ., value_col = sample)$p_value
  ) %>%
  pivot_longer(
    everything(),
    names_to = "test",
    values_to = "p_value"
  ) %>%
  convert_codes_to_factor(
    code_col = test,
    lookup_tbl = normality_tests_lookup_tbl,
    lookup_code_col = code,
    lookup_label_col = label,
    new_factor_col_name = test_factor
  ) %>%
  select(test_factor, p_value)
```

The results of the normality tests are as follows: 
```{r}
#| echo: false
normality_test_gamma_sample_tbl
```

\bigpause

As we can see, none of these tests was particularly successful, which is not surprising given that the gamma distribution is not normal. 

## Transformations 

### Transformations

To make our data more normal-like, we can apply various transformations to any given sample. Some of the most common transformations include:

\begin{enumerate}
\item \alert{Box-Cox transformation}: This is a family of power transformations that can be applied only to data that is always positive. It is given by the formula:
$$
\tilde{X} = \begin{cases}
\frac{X^\lambda - 1}{\lambda} & \text{if } \lambda \neq 0 \\
\log(X) & \text{if } \lambda = 0 
\end{cases}
$$
where $\lambda$ is some parameter.
\seti 

\end{enumerate} 

### 

\begin{enumerate} 

\conti

\item \alert{Yeo-Johnson transformation}: This transformation is similar to the Box-Cox transformation but can handle negative values. It is useful for data that is positively or negatively skewed.
It is given by the formula:
$$
\tilde{X} = \begin{cases}
\frac{((X + 1)^\lambda - 1)}{\lambda} & \text{if } X \geq 0, \lambda \neq 0 \\
\frac{-((-X + 1)^{2 - \lambda} - 1)}{2 - \lambda} & \text{if } X < 0, \lambda \neq 2 \\
\log(X + 1) & \text{if } X \geq 0, \lambda = 0 \\
\log(-X + 1) & \text{if } X < 0, \lambda = 2 
\end{cases}
$$
where $\lambda$ is some parameter. 
\end{enumerate}


We will now apply the Box-Cox transformation to the gamma sample data with parameters $\lambda = 0$, $0.25$, $0.5$, $0.75$, and $1$:

###

```{r}
#| fig-cap: "Comparison of different Box-Cox transformations."
#| fig-width: 14
#| fig-height: 7
#| echo: false

box_cox_lookup_tbl <- tibble(
  code = c("box_cox_0", "box_cox_025", "box_cox_05", "box_cox_075", "box_cox_1"),
  label = c(
    "lambda = 0",
    "lambda = 0.25",
    "lambda = 0.5",
    "lambda = 0.75",
    "lambda = 1"
  ),
)

box_cox_gamma_sample_tbl <- gamma_sample_tbl %>%
  mutate(
    box_cox_0 = bcPower(sample, lambda = 0),
    box_cox_025 = bcPower(sample, lambda = 0.25),
    box_cox_05 = bcPower(sample, lambda = 0.5),
    box_cox_075 = bcPower(sample, lambda = 0.75),
    box_cox_1 = bcPower(sample, lambda = 1)
  ) %>%
  pivot_longer(
    starts_with("box_cox_"),
    names_to = "lambda",
    values_to = "transformed_sample"
  ) %>%
  convert_codes_to_factor(
    code_col = lambda,
    lookup_tbl = box_cox_lookup_tbl,
    lookup_code_col = code,
    lookup_label_col = label,
    new_factor_col_name = lambda_factor
  )


box_cox_gamma_qq_plot <- box_cox_gamma_sample_tbl %>%
  ggplot(aes(sample = transformed_sample)) +
  facet_wrap(
    ~lambda_factor,
    nrow = 5,
    scales = "free"
  ) +
  geom_qq(
    color = c_palette["C red"],
    size = 0.3
  ) +
  geom_qq_line(
    color = c_palette["C blue"],
    linewidth = 0.5
  ) +
  labs(
    title = "QQ Plots",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_krul()

box_cox_gamma_histogram <- box_cox_gamma_sample_tbl %>%
  ggplot(aes(x = transformed_sample)) +
  facet_wrap(
    ~lambda_factor,
    nrow = 5,
    scales = "free"
  ) +
  geom_histogram(
    bins = 30,
    fill = c_palette["C blue"],
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Histograms",
    x = "Sample Values",
    y = "Density"
  ) +
  theme_krul()

box_cox_gamma_plot <- box_cox_gamma_qq_plot + box_cox_gamma_histogram +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Comparison of Box-Cox Transformations with Different Lambda Values",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )

box_cox_gamma_plot
```

## Normality Score Optimization

### Normality Score Optimization

We wish to find the optimal value of $\lambda$ that maximizes the log-likelihood of the transformed data. 


\bigpause

The following figure shows the log-likelihood profile for the Box-Cox transformation applied to the gamma sample data:


###
```{r}
#| echo: false
#| fig-cap: "Log-likelihood profile for the Box-Cox transformation."
#| fig-width: 14
#| fig-height: 6

# Intercept-only model
fit <- lm(sample ~ 1, data = gamma_sample_tbl)

# Estimate lambda
bc <- powerTransform(fit)

# Extract log-likelihood profile
bc_prof <- boxCox(fit, lambda = seq(-2, 2, 0.1), plotit = FALSE)

# Convert to tibble for ggplot
df <- tibble(lambda = bc_prof$x, loglik = bc_prof$y)

ggplot(df, aes(x = lambda, y = loglik)) +
  geom_line(color = c_pal("C blue"), linewidth = 1) +
  geom_vline(xintercept = bc$lambda, linetype = "dashed", color = c_pal("C red")) +
  labs(
    title = "Box–Cox Transformation",
    x = expression(lambda),
    y = "Log-likelihood"
  ) +
  theme_krul()
```

### 

Finally, we can extract the optimal value of $\lambda$ and transform the data using this value:
```{r}
#| echo: false
optimal_lambda <- bc$lambda

optimal_gamma_sample_tbl <- gamma_sample_tbl %>%
  mutate(
    transformed_sample = bcPower(sample, lambda = optimal_lambda)
  )

optimal_gamma_sample_tbl %>%
  summarise(
    shapiro = tidy_shapiro_test(data = ., value_col = transformed_sample)$p_value,
    ks = tidy_ks_test(data = ., value_col = transformed_sample)$p_value,
    ad = tidy_ad_test(data = ., value_col = transformed_sample)$p_value,
    jb = tidy_jb_test(data = ., value_col = transformed_sample)$p_value
  ) %>%
  pivot_longer(
    everything(),
    names_to = "test",
    values_to = "p_value"
  ) %>%
  convert_codes_to_factor(
    code_col = test,
    lookup_tbl = normality_tests_lookup_tbl,
    lookup_code_col = code,
    lookup_label_col = label,
    new_factor_col_name = test_factor
  ) %>%
  select(test_factor, p_value)
```
\bigpause 

As we can see, the Box-Cox transformation with the optimal value of $\lambda$ significantly improves the normality of the data, as all the normality tests now yield high $p$-values. We now show the QQ plot and histogram of the transformed data:

###
```{r}
#| fig-cap: "QQ plot and histogram of the Optimal Gamma Distribution Transformation."
#| fig-width: 10
#| fig-height: 5
#| echo: false

optimal_gamma_qq_plot <- optimal_gamma_sample_tbl %>%
  ggplot(aes(sample = transformed_sample)) +
  geom_qq(
    color = c_palette["C red"],
    size = 0.3
  ) +
  geom_qq_line(
    color = c_palette["C blue"],
    linewidth = 0.5
  ) +
  labs(
    title = "QQ Plot",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_krul()

optimal_gamma_histogram <- optimal_gamma_sample_tbl %>%
  ggplot(aes(x = transformed_sample)) +
  geom_histogram(
    bins = 30,
    fill = c_palette["C blue"],
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Histogram",
    x = "Sample Values",
    y = "Density"
  ) +
  theme_krul()

optimal_gamma_plot <- optimal_gamma_qq_plot + optimal_gamma_histogram +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Optimal Gamma Distribution Transformation",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )

optimal_gamma_plot
```
