# Introducción al análisis multivariante 

### Introducción al análisis multivariante

```{r}
#| output: false
#| echo: false
library(tidyverse)
library(broom)
library(here)
library(patchwork)
library(krulRutils)
library(magrittr)
library(e1071) # for skewness and kurtosis functions
library(car)
library(ggforce)
options(scipen = 999)
```


En esta presentación introduciremos conceptos básicos del análisis multivariante, enfocándonos en el vector de medias y la matriz de covarianzas de un vector aleatorio.

\bigpause

Estos conceptos son fundamentales para entender las relaciones entre múltiples variables en un conjunto de datos.

## El vector de medias y la matriz de covarianzas

### El vector de medias y la matriz de covarianzas

\begin{definition}
Let $\Pi$ be a population and let 
$$ 
X: \Pi \to \mathbb{R}^{p} 
$$ 
sea un vector aleatorio (renglón). El \alert{vector de medias} de $X$ se define como
$$
\mu = \mathbb{E}[X] = \left[ \mathbb{E}[X_1], \mathbb{E}[X_2], \ldots, \mathbb{E}[X_p] \right] = [\mu_1, \mu_2, \ldots, \mu_p]
$$
donde $X_i$ es la $i$‑ésima componente de $X$.
La \alert{matriz de covarianzas} de $X$ se define como 
$$
\Sigma = \Cov[X] = \mathbb{E}[(X - \mu)^{T}(X - \mu)] = \left[ \Cov[X_i, X_j] \right]_{i,j=1}^{p}
$$
donde $\Cov[X_i, X_j] = \mathbb{E}[(X_i - \mu_{i})(X_j - \mu_{j})]$ es la covarianza entre las componentes $i$ y $j$.
\end{definition}


###

\begin{remark}
Recuerde que $\Cov[X_i, X_j] = \Cov[X_j, X_i]$, por lo que la matriz es simétrica. Además, $\Cov[X_i, X_i] = \Var[X_i]$, la varianza de la componente $i$. Por ello también se le llama \alert{matriz varianza‑covarianza}.
\end{remark}

###

\begin{definition}
Dada una muestra de tamaño $n$ de $\Pi$, con matriz muestral $x$, definimos el \alert{vector de medias muestral} como 
$$
\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_i = \left[ \frac{1}{n} \sum_{i=1}^{n} x_{i,1}, \frac{1}{n} \sum_{i=1}^{n} x_{i,2}, \ldots, \frac{1}{n} \sum_{i=1}^{n} x_{i,p} \right]
$$
donde $x_i$ es el $i$‑ésimo renglón de $x$ y $x_{i,j}$ su componente $j$.
La \alert{matriz de covarianzas muestral} se define como 
$$
\hat{\Sigma} = \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^{T}(x_i - \bar{x})
$$
donde $\hat{\Sigma}$ es de tamaño $p \times p$ y $\hat{\Sigma}_{i,j} = \frac{1}{n-1} \sum_{k=1}^{n} (x_{k,i} - \bar{x}_i)(x_{k,j} - \bar{x}_j)$ es la covarianza muestral entre las componentes $i$ y $j$.
\end{definition}

###

\begin{remark}
La matriz de covarianzas muestral es un estimador insesgado de la matriz de covarianzas del vector aleatorio $X$.
\end{remark}


###

\begin{example}
Consider the following sample matrix of size $n=5$ from a population $\Pi$:
$$
\left[
\begin{array}{ccc}
2 & 5 & 7 \\
4 & 3 & 6 \\
6 & 4 & 5 \\
8 & 2 & 4 \\
10 & 1 & 3
\end{array}
\right]
$$
\end{example}

###

\begin{example}
To obtain the sample mean vector, we compute:
$$
\begin{aligned}
\bar{x}_1 &= \frac{2+4+6+8+10}{5} = 6 \\
\bar{x}_2 &= \frac{5+3+4+2+1}{5} = 3 \\
\bar{x}_3 &= \frac{7+6+5+4+3}{5} = 5
\end{aligned}
$$

\pause

In other words,
$$
\bar{x} = [6, 3, 5]
$$
\end{example}

###

\begin{example}
The sample variance of $X_j$ is:
$$
\hat{\Sigma}_{X_j} = \frac{1}{n-1} \sum_{i=1}^n (x_{ij} - \bar{x_j})^2
$$

\pause

Therefore,
$$
\begin{aligned}
\hat{\Sigma}_{X_1} &= \frac{(2-6)^2 + (4-6)^2 + (6-6)^2 + (8-6)^2 + (10-6)^2}{4} = 10 \\
\hat{\Sigma}_{X_2} &= \frac{(5-3)^2 + (3-3)^2 + (4-3)^2 + (2-3)^2 + (1-3)^2}{4} = 2.5 \\
\hat{\Sigma}_{X_3} &= \frac{(7-5)^2 + (6-5)^2 + (5-5)^2 + (4-5)^2 + (3-5)^2}{4} = 2.5
\end{aligned}
$$

\end{example}

###

\begin{example}
The sample covariance between $X_j$ and $X_k$ is:
$$
\hat{\Sigma}_{X_j, X_k} = \frac{1}{n-1} \sum_{i=1}^n (x_{ij} - \bar{x}_j)(x_{ik} - \bar{x}_k)
$$
This computations can be done in a similar way as before, the results are:
$$
\begin{aligned}
\hat{\Sigma}_{X_1, X_2}  = -4.5 \\
\hat{\Sigma}_{X_1, X_3}  = -5 \\
\hat{\Sigma}_{X_2, X_3}  = 2.25
\end{aligned}
$$

\end{example}

###

\begin{example}

The resulting covariance matrix is then:
$$
\hat{\Sigma}=
\left[
\begin{array}{ccc}
10 & -4.5 & -5 \\
-4.5 & 2.5 & 2.25 \\
-5 & 2.25 & 2.5
\end{array}
\right]
$$
\end{example}

## La distancia de Mahalanobis 

### La distancia de Mahalanobis

\begin{definition}
Suponga que tenemos una población $\Pi$ con un vector aleatorio $X$ de media $\mu$ y covarianzas $\Sigma$. Si $\Sigma$ es invertible, la \alert{distancia de Mahalanobis} entre $x$ e $y$ en $\mathbb{R}^p$ se define como:
$$
d_M(x, y) = \sqrt{(x - y) \Sigma^{-1} (x - y)^T}
$$
\end{definition}

###

En el contexto muestral, puede calcularse usando la matriz de covarianzas muestral $\hat{\Sigma}$:

\pause

\begin{definition}
Dada una muestra de tamaño $n$ de $\Pi$, con matriz muestral $x$ y matriz $\hat{\Sigma}$, la \alert{distancia de Mahalanobis muestral} entre $x_i$ y $x_j$ es:
$$
d_M(x_i, x_j) = \sqrt{(x_i - x_j) \hat{\Sigma}^{-1} (x_i - x_j)^{T}}
$$
\end{definition}

\pause

\begin{remark}
La distancia de Mahalanobis mide la separación en un espacio multivariante considerando las correlaciones entre variables. Es útil para identificar atípicos.
\end{remark}

###

To illustrate this point, we will consider the following dataset:

\bigpause 

We start by taking a random sample $x$ of size $n=300$ from a uniform distribution in the interval $[0, 10]$.
```{r}
#| echo: false
set.seed(123)
n <- 300
x_data <- runif(n, min = 0, max = 10)
``` 
\pause 

We will construct some error vector $\epsilon$ from a normal distribution with mean 0 and standard deviation 1.
```{r}
#| echo: false
epsilon_data <- rnorm(n, mean = 0, sd = 1)
```

\pause

We can now construct the vector
$$
y = x + \epsilon
$$
```{r}
#| echo: false
y_data <- x_data + epsilon_data

sample_data_tbl <- tibble(
  x = x_data,
  y = y_data
)
```
\pause

The associated scatter plot of the data is shown below:


###
```{r}
#| fig-cap: "Scatter plot of the sample data. Notice that the X and Y variables are highly correlated."
#| fig-width:  5
#| fig-height: 5
#| echo: false
sample_data_plot <- sample_data_tbl %>%
  ggplot(aes(x = x_data, y = y_data)) +
  geom_point(
    color = "black",
    size = 0.2,
    alpha = 1
  ) +
  coord_fixed() +
  labs(
    title = "",
    x = "",
    y = ""
  ) +
  theme_krul()

sample_data_plot
```

###
We now add the points (1, 1) and (5, 2) to our data:


###
```{r}
#| fig-cap: "Scatter Plot with extra sample points. Notice that one of these points is clearly an outlier."
#| fig-width:  5
#| fig-height: 5
#| echo: false
extra_points_tbl <- tibble(
  x_extra = c(1, 5),
  y_extra = c(1, 2)
)

extra_sample_data_plot <- sample_data_plot +
  geom_point(
    data = extra_points_tbl,
    aes(x = x_extra, y = y_extra),
    color = c_pal("C red"),
    size = 1.5
  )

extra_sample_data_plot
```

###
Notice that one of these points is clearly an outlier.

\bigpause 

We would like to have some concrete measure to show this.

\bigpause

```{r}
#| echo: false
sample_mean_vector <- colMeans(sample_data_tbl)
sample_covariance_matrix <- cov(sample_data_tbl)

sample_mean_tbl <- tibble(
  x_bar = sample_mean_vector[1],
  y_bar = sample_mean_vector[2]
)
```
However, if we use the Euclidean distance, we obtain that it is actually the outlier point that is closer to the sample mean vector:

###
```{r}
#| fig-cap: "Scatter Plot with sample mean vector and a circle representing points that are at a constant Euclidean distance from the center of our data."
#| fig-width:  5
#| fig-height: 5
#| echo: false
euclidean_sample_data_plot <- extra_sample_data_plot +
  geom_point(
    data = sample_mean_tbl,
    aes(x = x_bar, y = y_bar),
    color = c_pal("C red"),
    size = 1.5
  ) +
  geom_circle(
    aes(x0 = sample_mean_vector[1], y0 = sample_mean_vector[2], r = 4),
    color = c_pal("C blue"),
    linewidth = 0.6,
    fill = NA
  )

euclidean_sample_data_plot
```


###
On the other hand, if we compute the Mahalanobis distance, we obtain that the outlier point is actually further away from the sample mean vector:


###
```{r}
#| fig-cap: "Scatter Plot with sample mean vector and a ellipse representing points that are at a constant Mahalanobis distance from the center of our data."
#| fig-width:  5
#| fig-height: 5
#| echo: false
annotated_sample_data_plot <- extra_sample_data_plot +
  geom_point(
    data = sample_mean_tbl,
    aes(x = x_bar, y = y_bar),
    color = c_pal("C red"),
    size = 1.5
  ) +
  stat_ellipse(type = "norm", level = 0.8647, color = c_pal("C blue"), linewidth = 0.6)


annotated_sample_data_plot
```

## Scaling and Standardization 


### Scaling and Standardization

Now assume that we have a random vector $X$ with mean vector $\mu$ and covariance matrix $\Sigma$. We can define a new random vector $Y$ as follows: 
$$
Y^{T} = A X^{T} + b 
$$
where $A$ is a $p \times p$ matrix and $b$ is a $p$-dimensional vector.

\bigpause 

The mean vector and covariance matrix of the new random vector $Y$ can be computed as follows:
$$
\mu_Y^{T} = \mathbb{E}[Y]^{T} = A \mu^{T} + b 
$$
and
$$
\Sigma_Y = \Cov[Y] = A \Sigma A^{T} 
$$

###

As we will show later, it is always possible to find a matrix $A$ and a vector $b$ such that the new random vector $Y$ has mean vector $\mu_Y = 0$ and covariance matrix $\Sigma_Y = I_p$, where $I_p$ is the identity matrix of size $p \times p$.

\bigpause

This process is known as a \alert{standardization} of the random vector $X$.

## Activity: The Geometry of Multivariate Data

### Activity: The Geometry of Multivariate Data

\textbf{Part A: Theoretical Exercises}

\begin{enumerate}
    \item Prove that the Mahalanobis distance reduces to the Euclidean distance if the covariance matrix is the identity.
    \item Show that Mahalanobis distance is invariant under affine transformations.
    \item Show that if you standardize each variable in a multivariate dataset, the covariance matrix becomes the coorrelation matrix.
    \item Assume that $X_1$ and $X_2$ are two independent random variables following the standard normal distribution. Show that $d_M(X, 0)^{2}$ follows a chi-square distribution with 2 degrees of freedom, where $d_M(X, 0)$ is the Mahalanobis distance from the origin.
\end{enumerate}

###

\textbf{Part B: Practical Exercises}

\begin{enumerate}
    \item Generate a 2D dataset of 50 points (two correlated variables).
    \item Compute the sample mean vector and covariance matrix.
    \item Write a function in R that computes the Mahalanobis distance of each point from the mean.
    \item Identify the point with the largest Mahalanobis distance.
    \item Plot the dataset using \texttt{ggplot2}.
    \item Draw Mahalanobis distance ellipses for distances 1, 2, and 3.
    \item Color points depending on whether their distance is greater than 1, 2, or 3.
    \item Create a 3-variable dataset.
    \item Compute Mahalanobis distances in 3D.
    \item Plot pairs of variables with projected ellipses.
    \item Compare Euclidean distances from the mean with Mahalanobis distances.
\end{enumerate}


###

\textbf{Part C: Advanced Exercises}

\begin{enumerate}
    \item Simulate 5 points in 3 dimensions with one variable as a linear combination of the others.
    \item Try to compute Mahalanobis distance and discuss singularity of covariance.
    \item Apply a linear transformation to a 2D dataset and verify invariance of Mahalanobis distances.
    \item Write an R function that returns points forming a Mahalanobis-distance ellipse.
    \item Overlay Mahalanobis-distance ellipses of arbitrary radius on a scatter plot.
\end{enumerate}
