# Introducción al Tidyverse

# Los momentos estandarizados

# Pruebas de normalidad

# Introducción al análisis multivariante


# Distribución normal multivariante



## Definición y propiedades básicas

```{r}
#| output: false
#| echo: false
library(tidyverse)
library(broom)
library(here)
library(patchwork)
library(krulRutils)
library(magrittr)
library(e1071) # for skewness and kurtosis functions
library(car)
library(ggforce)
library(mvtnorm)
library(MVN)
library(ellipse)

options(scipen = 999)
```

### Definición y propiedades básicas

\begin{definition}
Sea $\mu$ un vector (renglón) y $\Sigma$ una matriz simétrica definida positiva. Decimos que un vector aleatorio (renglón) $X$ tiene \alert{distribución normal multivariante} con media $\mu$ y matriz de covarianzas $\Sigma$, en símbolos $X \sim \mathcal{N}_{p}(\mu, \Sigma)$, si su densidad es
$$
\frac{1}{(2\pi)^{p/2} |\Sigma|^{1/2}} \exp\left(-\frac{1}{2}(X - \mu) \Sigma^{-1} (X - \mu)^{T}\right)dX
$$
\end{definition}


### 
\begin{remark} Algunas observaciones sobre la definición:
\begin{enumerate}
\item Al igual que en el caso univariado, la normal multivariante queda completamente determinada por $\mu$ y $\Sigma$. \pause
\item La matriz $\Sigma$ debe ser simétrica y definida positiva para que la distribución esté bien definida y el exponente sea negativo definido. \pause
\item El término 
$$
|\Sigma| = \det(\Sigma)
$$
es la \alert{varianza generalizada} de la distribución. Puede interpretarse como una medida de la "dispersión" en el espacio de dimensión $p$. \pause
\item El término $(X - \mu) \Sigma^{-1} (X - \mu)^{T}$ es el cuadrado de la distancia de Mahalanobis entre $X$ y la media $\mu$. 
\end{enumerate}
\end{remark}

###

Como en el caso univariado, la normal multivariante es destacada por el Teorema Central del Límite, que a grandes rasgos dice que la suma de muchas variables i.i.d. (debidamente estandarizadas) converge a una normal. 

### Teorema central del límite (multivariante)

\begin{theorem}
Let $\{X_i\}_{i=1}^{\infty}$ be a sequence of independent and identically distributed 
random vectors in $\mathbb{R}^p$ with mean vector $\mu$ and covariance matrix $\Sigma$. 
Define the sample mean
\[
\bar{X}_n = \frac{1}{n}\sum_{i=1}^n X_i.
\]
Then, as $n \to \infty$,
\[
\sqrt{n}\,(\bar{X}_n - \mu) \;\xrightarrow{d}\; \mathcal{N}_p(0, \Sigma),
\]
where $\xrightarrow{d}$ denotes convergence in distribution and 
$\mathcal{N}_p(0, \Sigma)$ is the $p$-dimensional multivariate normal distribution 
with mean vector $0$ and covariance matrix $\Sigma$.
\end{theorem}

###

\begin{remark}
Even though the Central Limit Theorem singles out the normal multivariate distribution as a particularly important distribution, it is not the only relevant distribution in multivariate statistics.

\bigpause 

Actually, most real-world data does not follow a normal distribution, and many statistical methods are designed to handle non-normal data.

\bigpause

Furthermore, important statistical methods, like linear regression, principal component analysis and clustering do not yield good results when applied to a normal multivariate distribution.
\end{remark}

## Contornos de la normal multivariante


### Contornos de la normal multivariante
Observe que la función de densidad de la normal multivariante es gaussiana en $p$ dimensiones.

\bigpause 

Esto implica que los contornos son elipsoides centrados en $\mu$ y cuya forma está controlada por $\Sigma$.

\bigpause

La probabilidad asociada a estos contornos está controlada por la distribución $\chi^2$ en el siguiente sentido:

###

Let $0 \leq \varpi \leq 1$ be a probability level, and let $c$ be such that 
$$
\mathbb{P}(\chi^2_{p} \leq c) = \varpi
$$

\bigpause 

Entonces, el contorno que contiene una proporción $\varpi$ de la masa de probabilidad está dado por los puntos $X$ que satisfacen
$$
(X - \mu) \Sigma^{-1} (X - \mu)^{T} = c 
$$

###

In the following figure, we show the contours of the bivariate normal distribution with different covariance matrices.

\bigpause

The contours are ellipses centered at the mean $\mu = (0, 0)$, and their shape is determined by the covariance matrix $\Sigma$.

\bigpause

The color gradient represents the density of the distribution, with darker colors indicating higher density.


###
```{r}
#| fig-cap: "Distribución normal bivariante con distintas matrices de covarianzas"
#| fig-width: 12
#| fig-height: 6
#| echo: false

# Add seed for reproducibility
set.seed(123)

# Parameters
mu <- c(0, 0)
Sigma_1 <- matrix(c(1, 0, 0, 1), nrow = 2)
Sigma_2 <- matrix(c(0.3, 0, 0, 1.5), nrow = 2)
Sigma_3 <- matrix(c(1, 0.6, 0.6, 1), nrow = 2)
Sigma_4 <- matrix(c(1, -0.6, -0.6, 1), nrow = 2)



# Function to generate contour + scatter plot
plot_bivariate_normal <- function(mu, Sigma, n_samples = 1000,
                                  xlim = c(-3, 3), ylim = c(-3, 3),
                                  grid_length = 100) {
  # Create grid
  x <- seq(xlim[1], xlim[2], length.out = grid_length)
  y <- seq(ylim[1], ylim[2], length.out = grid_length)
  grid <- expand.grid(x = x, y = y)
  grid$z <- dmvnorm(grid, mean = mu, sigma = Sigma)

  # Sample points
  samples <- rmvnorm(n_samples, mean = mu, sigma = Sigma)
  colnames(samples) <- c("x", "y")
  samples <- as_tibble(samples)

  # Plot
  normal_plot <- ggplot() +
    geom_contour(data = grid, aes(x, y, z = z, color = after_stat(level))) +
    geom_point(
      data = samples,
      aes(x, y),
      color = c_pal("C red"),
      alpha = 0.5,
      size = 0.1
    ) +
    scale_color_viridis_c(option = "mako") +
    coord_cartesian(xlim = xlim, ylim = ylim) +
    labs(
      title = "",
      x = expression(X[1]), y = expression(X[2]),
      color = "Densidad"
    ) +
    theme_krul()

  return(normal_plot)
}

normal_plot_1 <- plot_bivariate_normal(mu, Sigma_1)
normal_plot_2 <- plot_bivariate_normal(mu, Sigma_2)
normal_plot_3 <- plot_bivariate_normal(mu, Sigma_3)
normal_plot_4 <- plot_bivariate_normal(mu, Sigma_4)

normal_plot <- (normal_plot_1 + normal_plot_2) /
  (normal_plot_3 + normal_plot_4) +
  plot_layout(widths = c(1, 1), heights = c(1, 1)) +
  plot_annotation(
    title = "Contornos de la normal bivariante ",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )

normal_plot
```

## Estandarización de la normal multivariante 

### Standardization of the Normal Multivariate Distribution

Just like in the univariate case, given a random vector $X \sim \mathcal{N}_{p}(\mu, \Sigma)$, we can standardize it to obtain a new random vector $Z$ that follows the standard multivariate normal distribution $\mathcal{N}_{p}(0, I_p)$, where $I_p$ is the identity matrix of size $p$.

\bigpause 

To describe this standardization process, we first need to remember some linear algebra results. 

### Singular Value Decomposition

\begin{theorem}
Let $\Sigma$ be a symmetric positive definite matrix of size $p \times p$. Then, there exists an orthogonal matrix $Q$ and a diagonal matrix $\Lambda$ with positive entries such that 
$$
\Sigma = Q \Lambda Q^{T}
$$
\end{theorem}

\pause

\begin{remark}
Remember that a matrix $Q$ is orthogonal if $Q^{T}Q = I_p$, where $I_p$ is the identity matrix of size $p$. This means that the columns of $Q$ are orthonormal vectors, and thus they form a basis for $\mathbb{R}^p$.
\end{remark}

###

\begin{remark}
The singular value decomposition (SVD) is a powerful tool in linear algebra that allows us to decompose a matrix into its constituent parts.

\bigpause 

In the case of the covariance matrix $\Sigma$, the SVD provides a way to express it in terms of its \alert{singular values} (the diagonal entries of $\Lambda$, also known as its \alert{eigenvalues}) and its \alert{singular vectors} (the columns of $Q$, also known as its \alert{eigenvectors}).

\bigpause

This decomposition is particularly useful for understanding the geometry of the multivariate normal distribution.
\end{remark}

###

Notice that, since $\Sigma$ is positive definite, the diagonal entries of
$$
\Lambda = \left[
\begin{array}{cccc}
\lambda_1 & 0 & \cdots & 0 \\
0 & \lambda_2 & \cdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \cdots & \lambda_p
\end{array}
\right]
$$
are all positive.

\bigpause 

In particular, we can define a new matrix 
$$
D = \left[
\begin{array}{cccc}
\sqrt{\lambda_1} & 0 & \cdots & 0 \\
0 & \sqrt{\lambda_2} & \cdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \cdots & \sqrt{\lambda_p}
\end{array}
\right]
$$
satisfying $D^2 = \Lambda$.

###

If we now take
$$ 
\sigma = D Q^{T}
$$

\bigpause

then 
$$
\Sigma = \sigma^{T} \sigma
$$

\bigpause

In particular, we can define the \alert{standardization} of the random vector $X$ as 
$$
Z = (X - \mu) \sigma^{-1} = (X - \mu) Q D^{-1} = (X - \mu) Q \Lambda^{-1/2}
$$
in which case $Z \sim \mathcal{N}_{p}(0, I_p)$.

## Geometric Interpretation of the SVD

### Geometric Interpretation of the SVD

Earlier in this chapter we considered the Normal Multivariate Distribution with mean 
$$ 
\mu = [0, 0] 
$$ 
and covariance matrix 
$$
\Sigma = \left[
\begin{array}{cc}
1 & 0.6 \\
0.6 & 1 
\end{array} 
\right] 
$$

\bigpause 

According to the SVD, we can write this covariance matrix as 
$$ 
\Sigma = Q \Lambda Q^{T} 
$$
for some orthogonal matrix $Q$ and diagonal matrix $\Lambda$. 

###
```{r}
#| echo: false
# Example mean and covariance
mu <- c(0, 0)
Sigma <- matrix(c(
  1, 0.6,
  0.6, 1
), 2, 2)


# Compute eigenvectors and eigenvalues
eig <- eigen(Sigma)
vectors <- eig$vectors
values <- eig$values
``` 

In this case we obtain: 
$$
Q = \left[
\begin{array}{cc}
 `{r} vectors[1, 1]` & `{r} vectors[1, 2]` \\
 `{r} vectors[2, 1]` & `{r} vectors[2, 2]`
\end{array}
\right]
$$
and 
$$
\Lambda = \left[
\begin{array}{cc}
 `{r} values[1]` & 0 \\
0 & `{r} values[2]`
\end{array}
\right]
$$
where the singular vectors are the columns of $Q$ and the singular values are the diagonal entries of $\Lambda$.

\bigpause 

Geometrically, the singular vectors represent the directions of the axes of the ellipsoid that defines the contours of the normal multivariate distribution, and the square root of the singular values represent the lengths of these axes.

###
```{r}
#| fig-cap: ""
#| fig-width: 3.5
#| fig-height: 3.5
#| fig-align: center
#| echo: false


# Scale eigenvectors by sqrt(eigenvalues) for plotting
vec_data <- tibble(
  x0 = mu[1],
  y0 = mu[2],
  x1 = mu[1] + sqrt(values) * vectors[1, ],
  y1 = mu[2] + sqrt(values) * vectors[2, ]
)

# Generate points of the 1-Mahalanobis-distance ellipse
ellipse_points <- as_tibble(ellipse(Sigma, centre = mu, level = 0.393))
# level = 0.393 corresponds to ~1 standard deviation for 2D


# Plot
ggplot() +
  geom_path(
    data = ellipse_points,
    aes(x = x, y = y),
    color = c_pal("C blue"),
    linewidth = 1
  ) +
  geom_segment(
    data = vec_data,
    aes(x = x0, y = y0, xend = x1, yend = y1),
    arrow = arrow(length = unit(0.2, "cm")),
    color = c_pal("C green"),
    linewidth = 1
  ) +
  geom_point(
    aes(x = mu[1], y = mu[2]),
    color = c_pal("C red"),
    size = 3
  ) +
  coord_equal() +
  labs(
    title = "",
    x = "",
    y = ""
  ) +
  theme_krul()
```


## Closure under Linear Transformations 

### Closure under Linear Transformations

\begin{theorem}
Let $X \sim \mathcal{N}_{p}(\mu, \Sigma)$ be a random vector with normal multivariate distribution. Then, for any matrix $A$ of size $m \times p$ and any vector $b$ of size $m$, the random vector 
$$
Y = AX + b 
$$ 
also has a normal multivariate distribution, specifically $Y \sim \mathcal{N}_{m}(A\mu + b, A\Sigma A^{T})$. 
\end{theorem}


### 

\begin{remark} 
This theorem states that if we apply a linear transformation to a random vector with a normal multivariate distribution, the resulting random vector will also have a normal multivariate distribution.

\bigpause 

The mean and covariance of the transformed vector are given by the formulas in the theorem.

\bigpause

This property is particularly useful in applications such as regression analysis, where we often need to transform the data to fit a model while maintaining the normality of the residuals.

\bigpause

In practice, this theorem is often used to derive the distribution of linear combinations of random variables, which is a common operation in multivariate statistics. 

\end{remark}

###

\begin{example} 
Let $X \sim \mathcal{N}_{2}(\mu, \Sigma)$ be a random vector with mean $\mu = (1, 2)$ and covariance matrix 
$$ 
\Sigma = \left[ 
\begin{array}{cc} 
1 & 0.5 \\ 
0.5 & 2 
\end{array} 
\right] 
$$ 

\pause

Now, consider the linear transformation 
$$ 
Y = \begin{bmatrix} 2 & 1 \\ 0 & 3 \end{bmatrix} X + \begin{bmatrix} 0 \\ 1 \end{bmatrix} 
$$ 

\pause

In other words, 
\begin{align*}
Y_1 & = 2X_1 + X_2 \\ 
Y_2 & = 3X_2 + 1 
\end{align*}
\end{example} 

### 

\begin{example}

Using the closure under linear transformations theorem, we can find the mean and covariance of $Y$:
\begin{align*} 
\text{Mean: } & A\mu + b = \begin{bmatrix} 2 & 1 \\ 0 & 3 \end{bmatrix} \begin{bmatrix} 1 \\ 2 \end{bmatrix} + \begin{bmatrix} 0 \\ 1 \end{bmatrix} = \begin{bmatrix} 4 \\ 7 \end{bmatrix} \\ 
\text{Covariance: } & A\Sigma A^{T} = \begin{bmatrix} 2 & 1 \\ 0 & 3 \end{bmatrix} \begin{bmatrix} 1 & 0.5 \\ 0.5 & 2 \end{bmatrix} \begin{bmatrix} 2 & 0 \\ 1 & 3 \end{bmatrix} = \begin{bmatrix} 8 & 9 \\ 9 & 18 \end{bmatrix} 
\end{align*} 

\pause

Thus,  
$$Y \sim \mathcal{N}_{2}\left(\begin{bmatrix} 4 \\ 7 \end{bmatrix}, \begin{bmatrix} 8 & 9 \\ 9 & 18 \end{bmatrix}\right)
$$ 
\end{example}

###

\begin{remark} Notice that since projection into a subspace is a linear transformation, the closure under linear transformations theorem implies that any subset of the components of a normal multivariate distribution also has a normal multivariate distribution.

\bigpause

This is a very useful property, as it allows us to analyze subsets of variables while still maintaining the normality assumption. 
\end{remark}

## Multivariate Normality Testing 

### Multivariate Normality Testing

There are several statistical tests to check whether a given dataset follows a multivariate normal distribution. These tests are essential in multivariate statistics, as many methods assume normality. Here are some of the most commonly used tests, together with their implementation in R:

\begin{enumerate}
  \item \alert{Mardia's Test:}  
    Based on multivariate skewness and kurtosis.  
    For a multivariate normal distribution, skewness should be close to 0 and kurtosis close to \(p(p+2)\), where \(p\) is the number of variables. Implemented in R via \texttt{MVN::mvn(your\_data, mvn\_test = "mardia")}.
    
  \item \alert{Henze-Zirkler (HZ) Test:}  
    Uses the empirical characteristic function to measure deviations from normality. It has good global power and is suitable for moderate dimensions. Implemented in R via \texttt{MVN::mvn(your\_data, mvn\_test = "hz")}.
    
\seti 
\end{enumerate} 

### 
\begin{enumerate} 

\conti

  \item \alert{Royston's Test:}  
    A multivariate extension of the Shapiro-Wilk test, based on the product of univariate Shapiro-Wilk statistics applied to principal components. Easy to interpret, though power decreases in high dimensions. Implemented in R via \texttt{MVN::mvn(your\_data, mvn\_test = "royston")}.
    
  \item \alert{Doornik-Hansen Test:}  
    Combines skewness and kurtosis into a single omnibus test, providing a robust alternative to Mardia. Implemented in R via \texttt{MVN::mvn(your\_data, mvn\_test = "dh")}.
    
  \item \alert{Graphical Methods (Chi-square Q-Q plots):}  
    Plot the Mahalanobis distances of the data against the quantiles of a chi-square distribution. Deviations from a straight line indicate departures from multivariate normality. Implemented in R via \texttt{multivariate\_diagnostic\_plot(data = your\_data, type = "qq") }.
\end{enumerate}
