# An Introduction to the Tidyverse

# The Standardized Moments

# Normality Testing

# Introduction to Multivariate Analysis

# Normal Multivariate Distribution

# Principal Component Analysis

# Factor Analysis: Theoretical Models

# Factor Analysis: Practical Methods

# Hierarchical Clustering

### Hierarchical Clustering


```{r}
#| output: false
#| echo: false
library(tidyverse) # for data manipulation and visualization
library(broom) # for tidying data
library(here) # for file paths
library(patchwork) # for combining ggplots
library(krulRutils) # for custom package functions
library(magrittr) # for the pipe operator
library(e1071) # for skewness and kurtosis functions
library(car) # for robust scaling functions
library(ggforce) # for ggplot2 extensions
library(mvtnorm) # for multivariate normal distribution
library(MVN) # for multivariate normality tests
library(ellipse) # for generating covariance ellipses
library(ggrepel) # for better text annotations in ggplot2
library(lavaan) # for structural equation modeling


options(scipen = 999)
```


\alert{Clustering} is an unsupervised learning technique that groups similar data points together based on their features. 

\bigpause

Unlike supervised learning, clustering does not rely on predefined labels or categories.

\bigpause

Instead, it identifies patterns and structures within the data itself.

###

Clustering methods are usually categorized into two main families: \pause
\begin{enumerate} 
\item \alert{Hierarchical Clustering}: These methods create a \alert{dendrogram} (also known as a \alert{tree diagram}) representing the taxonomic structure of our data. It is specially useful for small datasets and when the number of clusters is unknown. \pause
\item \alert{Non-Hierarchical Clustering}: These methods, such as K-means and DBSCAN, partition the data into a predefined number of clusters or based on density. They are generally more scalable for larger datasets.
\end{enumerate}



## Understanding Dendrograms 

### Understanding Dendrograms

A dendrogram is a tree-like diagram that illustrates the taxonomic structure of our data according to a hierarchical clustering algorithm.

\bigpause

It provides a visual representation of how data points are grouped together based on their similarities.

\bigpause

The vertical axis represents the distance or dissimilarity between clusters, while the horizontal axis represents the individual data points or clusters.

###

By cutting the dendrogram at a specific height, we can determine the number of clusters that we want to consider for our analysis. 

\bigpause

As an illustration, lets consider the scatter plot of the Iris dataset (considering only the variables "Sepal.Length" and "Sepal.Width") where each point is colored according to the species of the flower:

###

```{r}
#| fig-cap: ""
#| fig-width: 9
#| fig-height: 5
#| echo: false

data(iris)

iris_plot <- iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(
    size = 0.8,
    alpha = 1
  ) +
  c_scale_color("C green", "C blue", "C red") +
  labs(
    title = "Iris Dataset: Sepal Length vs Sepal Width",
    x = "Sepal Length (cm)",
    y = "Sepal Width (cm)"
  ) +
  theme_krul() +
  theme(legend.position = "top")

iris_plot
```
###

We will now consider a dendrogram associated to this dataset and show how cutting it at different heights leads to different numbers of clusters: 

### 
```{r}
#| fig-cap: ""
#| fig-width:  9
#| fig-height: 5
#| echo: false

# Use the built-in iris dataset only
iris_tbl <- as_tibble(iris)

k_values <- c(1, 3, 4, 6)

# Cluster on the chosen 2D projection to match the plot
standardized_sepal_matrix <- iris_tbl %>%
  select(Sepal.Length, Sepal.Width) %>%
  scale()

distances <- stats::dist(standardized_sepal_matrix, method = "euclidean")
sepal_hclust <- stats::hclust(distances, method = "ward.D2")

sepal_tbl <- iris_tbl %>%
  select(Sepal.Length, Sepal.Width) %>%
  mutate(.row_id = row_number())

# Build long data for facets by k
long_pts <- lapply(k_values, function(k) {
  cluster <- stats::cutree(sepal_hclust, k = k)
  tibble::tibble(
    .row_id = seq_along(cluster),
    cluster = factor(as.integer(cluster)),
    k = factor(paste0("k = ", k), levels = paste0("k = ", k_values))
  )
})

long_pts <- bind_rows(long_pts)
sepal_cluster_tbl <- left_join(long_pts, sepal_tbl, by = ".row_id")

# Polygons for each (k, cluster)
sepal_poly_tbl <- cluster_polygons(
  sepal_cluster_tbl,
  Sepal.Length, Sepal.Width,
  k, cluster
)

sepal_cluster_plot <- sepal_cluster_tbl %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(color = "black", alpha = 1, size = 0.8) +
  labs(
    title = "Clusters",
    x = "Sepal Length",
    y = "Sepal Width"
  ) +
  theme_krul() +
  facet_wrap(~k, ncol = 1, scales = "fixed")

sepal_cluster_plot <- sepal_cluster_plot +
  geom_path(
    data = sepal_poly_tbl,
    aes(
      x = Sepal.Length,
      y = Sepal.Width,
      group = interaction(k, cluster),
      color = cluster
    ),
    linewidth = 0.8,
    alpha = 0.5
  ) +
  c_scale_color(
    "C green",
    "C blue",
    "C red",
    "C magenta",
    "C yellow",
    "C cyan"
  ) +
  labs(color = "Cluster")

# Dendrogram geometry (replicated across facets)
sepal_dendro_data <- ggdendro::dendro_data(sepal_hclust, type = "rectangle")
k_fac <- factor(paste0("k = ", k_values), levels = paste0("k = ", k_values))

dendro_segments_faceted <- sepal_dendro_data$segments %>%
  mutate(.rep = 1) %>%
  tidyr::crossing(k = k_fac) %>%
  select(-.rep)

# One dashed cut line per k facet
dendro_cuts_tbl <- tibble::tibble(
  k = k_fac,
  y_cut = vapply(
    k_values,
    function(k) cut_dendrogram_height_for_k(sepal_hclust, k),
    numeric(1)
  )
)

dendro_faceted <- ggplot() +
  geom_segment(
    data = dendro_segments_faceted,
    aes(x = x, y = y, xend = xend, yend = yend),
    linewidth = 0.4
  ) +
  geom_hline(
    data = dendro_cuts_tbl,
    aes(yintercept = y_cut),
    linetype = "dashed",
    color = c_pal("C red")
  ) +
  labs(title = "Dendrogram cuts", x = NULL, y = "Height") +
  theme_krul() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank()
  ) +
  facet_wrap(~k, ncol = 1, scales = "fixed")

combined_plot <-  (dendro_faceted | sepal_cluster_plot) +
  plot_layout(widths = c(1, 2))

combined_plot
```

## Performing Hierarchical Clustering

### Performing Hierarchical Clustering

There are several methods to perform hierarchical clustering, but the most common ones are: \pause
\begin{itemize}
\item \alert{Agglomerative Clustering}: This is a bottom-up approach where each data point starts as its own cluster, and pairs of clusters are merged iteratively based on their similarity until a single cluster is formed or a stopping criterion is met. \pause
\item \alert{Divisive Clustering}: This is a top-down approach where all data points start in a single cluster, and the cluster is recursively split into smaller clusters based on dissimilarity until each data point is its own cluster or a stopping criterion is met.
\end{itemize}

###

Once we have chosen the method, we need to define a distance metric to measure the similarity between data points.

\bigpause

Common distance metrics include: 

###
\begin{itemize} 
\item \alert{Euclidean Distance}: This is the most common distance metric, which calculates the straight-line distance between two points in a multi-dimensional space. \pause
\item \alert{Mahalanobis Distance}: This metric takes into account the correlations between variables and is useful when dealing with multivariate data. \pause
\item \alert{Canberra Distance}: This metric is sensitive to small changes in the data and is useful when dealing with sparse data. The formula for the Canberra distance between two points $p$ and $q$ is given by: 
$$ 
d(p, q) = \sum_{i=1}^{n} \frac{|p_i - q_i|}{|p_i| + |q_i|} 
$$
where $n$ is the number of dimensions, and $p_i$ and $q_i$ are the coordinates of points $p$ and $q$ respectively.
\end{itemize} 

### 

\begin{itemize}
\item \alert{Manhattan Distance}: Also known as the $L^{1}$ norm or taxicab distance, this metric calculates the distance between two points by summing the absolute differences of their coordinates. It is particularly useful when dealing with high-dimensional data or when the data has a grid-like structure. The formula for the Manhattan distance between two points $p$ and $q$ is given by: 
$$ 
d(p, q) = \sum_{i=1}^{n} |p_i - q_i| 
$$
\pause
\item \alert{Minkowski Distance}: This is a generalization of the Euclidean and Manhattan distances, which allows for different values of the parameter $p$ to control the distance calculation. The formula for the Minkowski distance between two points $p$ and $q$ is given by:
$$
d(p, q) = \left( \sum_{i=1}^{n} |p_i - q_i|^p \right)^{1/p} 
$$
When $p = 1$, it corresponds to the Manhattan distance, and when $p = 2$, it corresponds to the Euclidean distance.
\end{itemize}


###

Finally, we need to choose a linkage criterion to determine how the distance between clusters is calculated. Common linkage criteria include: \pause
\begin{itemize}
\item \alert{Single Linkage}: This criterion defines the distance between two clusters as the minimum distance between any pair of points from each cluster. It tends to create long, chain-like clusters and is sensitive to noise and outliers. \pause
\item \alert{Complete Linkage}: This criterion defines the distance between two clusters as the maximum distance between any pair of points from each cluster. It tends to create compact, spherical clusters and is less sensitive to noise and outliers compared to single linkage.
\end{itemize} 

###

\begin{itemize}
\item \alert{Average Linkage}: This criterion defines the distance between two clusters as the average distance between all pairs of points from each cluster. It provides a balance between single and complete linkage and is less sensitive to noise and outliers. \pause
\item \alert{Ward's Linkage}: This criterion minimizes the total within-cluster variance by merging clusters that result in the smallest increase in variance. It tends to create compact, spherical clusters and is less sensitive to noise and outliers compared to other linkage criteria. 
\end{itemize}

###

Once we have chosen the method, distance metric, and linkage criterion, we can perform hierarchical clustering using the `hclust` function in R.

\bigpause

As was shown in the previous example, we can then visualize the results using a dendrogram and cut the dendrogram at a specific height to determine the number of clusters.
