--- 
title: |
  Practice construction of confidence intervals,

  using the pacemaker dataset

format:
  pdf:
    pdf-engine: pdflatex
    fontsize: 11pt
    number-sections: true

output-dir: reports/

highlight-style: tango 

execute:
  echo: true  # Set to false if you don't want code displayed
---
In these notes, we will practice constructing confidence intervals in R using the pacemaker dataset. We will start by loading the necessary libraries and data.

```{r}
#| output: false
library(tidyverse)
library(krulRutils)
library(here)
library(GGally)
library(janitor)
library(patchwork)
library(broom)

pacemaker_tbl <- read_csv(
  here("data", "pacemaker.csv")
) %>%
  clean_names() %>%
  drop_na() %>%
  rename(
    period = periodo_entre_pulsos,
    intensity = intensidad_de_pulso,
  )
```
This dataset is supposed to contain information about a new pacemaker that is yet to hit the market. The purpose of this report is to analyse the information given and conclude whether the pacemaker provides any significant benefits to the patients. We start by constructing a lookup table for the "marcapasos" variable, and converting it into a factor variable.
```{r}
pacemaker_lookup_tbl <- tibble(
  code = c("Sin MP", "Con MP"),
  label = c("Without Pacemaker", "With Pacemaker")
)

pacemaker_factor <- pacemaker_tbl %>%
  convert_codes_to_factor(
    code_col = marcapasos,
    lookup_tbl = pacemaker_lookup_tbl,
    lookup_code_col = code,
    lookup_label_col = label,
    new_factor_col_name = pacemaker
  ) %>%
  select(-marcapasos)
```

# Analysis of the "period" variable
We will start by focusing on the "period" variable, which represents the time between pulses in milliseconds. 

## Histogram of the "period" variable
To start our analysis, we will create a histogram of the "period" variable for both groups of patients: those with a pacemaker and those without. This will help us visualize the distribution of the "period" variable in each group.
```{r}
#| fig-width: 12
#| fig-height: 6
#| fig-cap: "\\small Histogram of the period variable for both groups of patients"
pacemaker_period_histogram <- pacemaker_factor %>%
  ggplot(aes(x = period, fill = pacemaker)) +
  facet_wrap(~ pacemaker, ncol = 1) +
  geom_histogram(
    bins = 30,
    color = "black",
    alpha = 1
  ) +
  c_scale_fill("C blue", "C violet") +
  labs(
    title = "Histograms of Period Variable",
    x = "Period (ms)",
    y = "Count",
    fill = "Pacemaker"
  ) +
  theme_krul() +
  theme(
    legend.position = "none"
  )

pacemaker_period_histogram
```
As we can see, the patients without the pacemaker have a more uneven distribution of the "period" variable, while the patients with the pacemaker have a more normal distribution. In particular, there are quite a number of outliers in the "Without Pacemaker" group, which might not be beneficial for the patients. We will check this by looking at the boxplots of the "period" variable for both groups of patients.

## Boxplot of the "period" variable
As we mentioned, we will now create boxplots of the "period" variable for both groups of patients, this will help us visualize the distribution of this variable and identify any potential outliers.
```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "\\small Boxplot of the period variable for both groups of patients" 
pacemaker_period_boxplot <- pacemaker_factor %>%
  ggplot(aes(x = period, y = 0, fill = pacemaker, color = pacemaker)) +
  facet_wrap(~pacemaker, ncol = 1) +
  geom_boxplot(
    color = "black",
    outlier.shape = NA,
    alpha = 1
  ) +
  stat_boxplot(
    geom = "errorbar",
    color = "black",
    width = 0.3,
    linewidth = 0.5
  ) +
  geom_jitter(
    alpha = 0.8,
    size = 0.8,
    height = 0.3
  ) +
  c_scale_fill("C blue", "C violet") +
  c_scale_color("C rose", "C red") +
  labs(
    title = "Boxplots of Period Variable",
    x = "Period (ms)",
    y = ""
  ) +
  theme_krul() +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

pacemaker_period_boxplot
```
Notice that this plot confirms our previous observation that the "Without Pacemaker" group has a more uneven distribution of the "period" variable, with some outliers that might not be beneficial for the patients.


## Confidence intervals for the mean of the "period" variable
We will now construct the confidence intervals, with a confidence level of 0.95, for the mean of the "period" variable for both groups of patients. We start by creating a tibble containing all the relevant information, including the mean and the confidence intervals for each group.
```{r}
pacemaker_period_ci <- pacemaker_factor %>%
  group_by(pacemaker) %>%
  summarise(
    x_bar = mean(period),
    ci_lower = t.test(period, conf.level = 0.95)$conf.int[1],
    ci_upper = t.test(period, conf.level = 0.95)$conf.int[2],
  ) %>%
  ungroup()
```
With this information, we can now create a plot showing the confidence intervals for the mean of the "period" variable for both groups of patients.
```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "\\small Confidence intervals for the mean of the period variable for both groups of patients"
pacemaker_period_ci_plot <- pacemaker_period_ci %>%
  ggplot(aes(x = x_bar, y = pacemaker, xmin = ci_lower, xmax = ci_upper)) +
  geom_errorbar(
    width = 0.1,
    color = c_pal("C blue"),
    linewidth = 1
  ) +
  geom_point(
    size = 3,
    color = c_pal("C red")
  ) +
  labs(
    title = "Confidence Intervals for Mean of Period Variable",
    x = "Mean Period (ms)",
    y = ""
  ) +
  theme_krul()

pacemaker_period_ci_plot
```
From this plot, we can see that the confidence intervals for the mean of the "period" variable for both groups of patients do not overlap. This suggests that there is a significant difference between the two groups, and that the pacemaker does have an effect on the "period" variable. In particular, we can assert (wih 95% confidence) that the pacemaker indeed reduces the mean period between pulses.

# Analysis of the "intensity" variable 
We will now focus on the "intensity" variable, which represents the intensity of the pulse in milliamperes.

## Histogram of the "intensity" variable 

To start our analysis, we will create a histogram of the "intensity" variable for both groups of patients: those with a pacemaker and those without. This will help us visualize the distribution of the "intensity" variable in each group.
```{r}
#| fig-width: 12 
#| fig-height: 6 
#| fig-cap: "\\small Histogram of the intensity variable for both groups of patients"
pacemaker_intensity_histogram <- pacemaker_factor %>%
  ggplot(aes(x = intensity, fill = pacemaker)) +
  facet_wrap(~ pacemaker, ncol = 1) +
  geom_histogram(
    bins = 30,
    color = "black",
    alpha = 1
  ) +
  c_scale_fill("C blue", "C violet") +
  labs(
    title = "Histograms of Intensity Variable",
    x = "Intensity (mA)",
    y = "Count",
    fill = "Pacemaker"
  ) +
  theme_krul() +
  theme(
    legend.position = "none"
  )

pacemaker_intensity_histogram
```
As we can see, the patients with the pacemaker have a more uneven distribution of the "intensity" variable, while the patients without the pacemaker have a more normal distribution. In particular, there are quite a number of outliers in the "Without Pacemaker" group, which might not be beneficial for the patients. We will check this by looking at the boxplots of the "intensity" variable for both groups of patients. 

## Boxplot of the "intensity" variable 

As we mentioned, we will now create boxplots of the "intensity" variable for both groups of patients, this will help us visualize the distribution of this variable and identify any potential outliers.
```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "\\small Boxplot of the intensity variable for both groups of patients"
pacemaker_intensity_boxplot <- pacemaker_factor %>%
  ggplot(aes(x = intensity, y = 0, fill = pacemaker, color = pacemaker)) +
  facet_wrap(~pacemaker, ncol = 1) +
  geom_boxplot(
    color = "black",
    outlier.shape = NA,
    alpha = 1
  ) +
  stat_boxplot(
    geom = "errorbar",
    color = "black",
    width = 0.3,
    linewidth = 0.5
  ) +
  geom_jitter(
    alpha = 0.8,
    size = 0.8,
    height = 0.3
  ) +
  c_scale_fill("C blue", "C violet") +
  c_scale_color("C rose", "C red") +
  labs(
    title = "Boxplots of Intensity Variable",
    x = "Intensity (mA)",
    y = ""
  ) +
  theme_krul() +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

pacemaker_intensity_boxplot
```
Notice that this plot confirms our previous observation that the "Without Pacemaker" group has a more uneven distribution of the "intensity" variable, with some outliers that might not be beneficial for the patients.

## Confidence intervals for the mean of the "intensity" variable 
We will now construct the confidence intervals, with a confidence level of 0.95, for the mean of the "intensity" variable for both groups of patients. We start by creating a tibble containing all the relevant information, including the mean and the confidence intervals for each group.
```{r}
pacemaker_intensity_ci <- pacemaker_factor %>%
  group_by(pacemaker) %>%
  summarise(
    x_bar = mean(intensity),
    ci_lower = t.test(intensity, conf.level = 0.95)$conf.int[1],
    ci_upper = t.test(intensity, conf.level = 0.95)$conf.int[2],
  ) %>%
  ungroup()
```
With this information, we can now create a plot showing the confidence intervals for the mean of the "intensity" variable for both groups of patients.
```{r}
#| fig-width: 8 
#| fig-height: 6 
#| fig-cap: "\\small Confidence intervals for the mean of the intensity variable for both groups of patients"
pacemaker_intensity_ci_plot <- pacemaker_intensity_ci %>%
  ggplot(aes(x = x_bar, y = pacemaker, xmin = ci_lower, xmax = ci_upper)) +
  geom_errorbar(
    width = 0.1,
    color = c_pal("C blue"),
    linewidth = 1
  ) +
  geom_point(
    size = 3,
    color = c_pal("C red")
  ) +
  labs(
    title = "Confidence Intervals for Mean of Intensity Variable",
    x = "Mean Intensity (mA)",
    y = ""
  ) +
  theme_krul()

pacemaker_intensity_ci_plot
```
From this plot, we can see that the confidence intervals for the mean of the "intensity" variable for both groups of patients overlap quite a bit. This suggests that there is no significant difference between the two groups, and that the pacemaker does not have a significant effect on the "intensity" variable. 

# Conclusion

Based on the data collected, we can conclude that the pacemkaer indeed has a positive effect on the "period" variable with faster, more evenly distributed pulses. Its effect on the "intensity" variable, however, is not significant. We still recommend the use of the pacemaker, as it does provide a significant benefit to the patients in terms of the "period" variable, which is crucial for the proper functioning of the heart.




















