--- 
title: |
  An Introduction to Statistical Learning 
  
  with Applications in R 
  
  Chapter 3 Exercises

format:
  pdf:
    pdf-engine: pdflatex
    fontsize: 11pt

highlight-style: tango 

execute:
  echo: true  # Set to false if you don't want code displayed
---
We will start by loading the required libraries:
```{r}
#| output: false
library("tidyverse")
library("krulRutils")
library("here")
library("ISLR2")
library("cowplot")
library("rlang")
options(scipen = 999) # Disable scientific notation
```

# Simple Linear Regression

For the first part of this report, we will create a plot showing the relation between the variables contained in the "Advertising" dataset. We start by loading and cleaning the dataset:
```{r}
#| output: false
advertising_tbl <- read_csv(here("data", "Advertising.csv")) %>%
  select(-1)
```

We now create a function that generates a scatter plot for one of the explanatory variables against the response variable "sales". 
```{r}
#| output: false
make_advertising_plot <- function(var) {
  var_name <- enquo(var) |>
    as_name()
  advertising_tbl %>%
    mutate(
      fitted = fitted(lm(reformulate(var_name, response = "sales"), data = .))
    ) %>%
    ggplot(aes(x = {{var}}, y = sales)) +
    geom_smooth(method = "lm", se = TRUE, color = "#1f77b4") +
    geom_segment(
      aes(xend = {{var}}, yend = fitted),
      color = "#1f77b4",
      alpha = 0.5
    ) +
    geom_point(color = "#d62728", size = 1, alpha = 1) +
    theme(
      panel.grid.major = element_line(color = "grey80"),
      panel.grid.minor = element_line(color = "gray80")
    ) +
    labs(
      tittle = paste(str_to_title(var_name), "Advertising vs Sales"),
      x = paste(str_to_title(var_name), "Advertising Budget"),
      y = "Sales"
    )
}
```
With this function in place, we can now generate the individual plots for each explanatory variable:
```{r}
#| output: false
advertising_plot <- list(
  make_advertising_plot(TV),
  make_advertising_plot(radio),
  make_advertising_plot(newspaper)
)
```
We no put all of these plots together using "plot_grid" from the "cowplot" package:
```{r}
main_advertising_plot <- plot_grid(plotlist = advertising_plot, ncol = 3)
```
Finally, we display these plots:
```{r}
#| fig-width: 9
#| fig-height: 3
final_plot <- (ggdraw() +
  draw_label(
    "Comparative Analysis of Advertising Investement on 
    Tv, Radio, or Newspaper",
    fontface = "bold",
    x = 0.5,
    y = 0.93,
    hjust = 0.5,
    size = 12
  ) +
  draw_plot(main_advertising_plot, y = 0, height = 0.82)) %>%
  print()
```
  
