--- 
title: |
  Practice with multiple continous variables in R,

  Scatter Plots and Correlation

format:
  pdf:
    pdf-engine: pdflatex
    fontsize: 11pt
    number-sections: true

output-dir: reports/

highlight-style: tango 

execute:
  echo: true  # Set to false if you don't want code displayed
---

# Multiple Continuous Variables in R 

In this notes, we will practice with multiple continuous variables in R, focusing on scatter plots and correlation. We will start by loading the necessary libraries and the "airquality" dataset.

```{r}
#| output: false
library(tidyverse)
library(krulRutils)
library(GGally)
library(here)
options(scipen = 999) # Disable scientific notation
data(airquality)
```

We start by doing a simple cleaning of our dataset, removing rows with missing values and collecting the month and day into a single date column: 
```{r}
airquality_tbl <- airquality %>%
  as_tibble() %>%
  drop_na() %>%
  mutate(Date = as.Date(paste(1973, Month, Day, sep = "-"))) %>%
  select(-Month, -Day) %>%
  glimpse()
```

We now construct a matrix of scatter plots to visualize the relationships between the continuous variables in our dataset. The `ggpairs` function from the `GGally` package is particularly useful for this purpose: 
```{r}
#| fig-width: 12
#| fig-height: 12
#| fig-cap: "\\small Scatter plot matrix of the airquality dataset"
ggpairs_plot <- (airquality_tbl %>%
  select(-Date) %>%  # Exclude the Date column for scatter plots
  ggpairs() +
  labs(
    title = "Scatter Plot Matrix of Air Quality Data",
    subtitle = "Showing relationships between Ozone, Solar.R, Wind, and Temp"
  )
) %>%
  theme_krul_ggpairs()

ggsave(
  filename = here("images", "ggpairs_airquality.png"),
  plot = ggpairs_plot,
  width = 12,
  height = 12,
  dpi = 300
)

ggpairs_plot
```

We notice that the variables with the strongest correlation are "Ozone" and "Temperature". So we now focus on these two variables to create a more detailed scatter plot with a fitted regression line:
```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "\\small Scatter plot of Ozone vs Temperature with regression line"
ozone_temp_plot <- airquality_tbl %>%
  ggplot(aes(x = Temp, y = Ozone)) +
  geom_smooth(
    method = "lm",
    color = c_pal("C blue"),
    se = FALSE,
    linewidth = 0.5
  ) +
  geom_point(
    alpha = 1,
    color = c_pal("C red"),
    size = 1
  ) +
  labs(
    title = "Scatter Plot of Ozone vs Temperature",
    x = "Temperature (°F)",
    y = "Ozone (ppb)"
  ) +
  theme_krul()

ozone_temp_plot
```

Finally, we show a time series showing the temperature in New York City during the months of May to September 1973, which is the period covered by the "airquality" dataset:
```{r}
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "\\small Temperature in New York City from May to September 1973"
temp_time_series_plot <- airquality_tbl %>%
  ggplot(aes(x = Date, y = Temp)) +
  geom_line(color = c_pal("C blue"), linewidth = 0.5) +
  labs(
    title = "Temperature in New York City (1973)",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  theme_krul()

temp_time_series_plot
```
