# The Standardized Moments

As it is well known, any normal distribution is completely determined by its mean and standard deviation. However, many real-world datasets do not follow a normal distribution. In this chapter, we will explore some other distributions and show how the concepts of skewness and kurtosis can help us understand its shape. We will also introduce QQ plots, which are useful graphical tools for comparing the distribution of a dataset against a theoretical distribution, which is usually taken to be normal. But before starting, let's load the necessary libraries and set some options. 

```{r}
#| output: false
library(tidyverse)   # for data manipulation and visualization
library(broom)       # for tidying data
library(here)        # for file paths
library(patchwork)   # for combining ggplots
library(krulRutils)  # for custom package functions
library(magrittr)    # for the pipe operator
library(e1071)       # for skewness and kurtosis functions
library(car)         # for robust scaling functions
library(ggforce)     # for ggplot2 extensions
library(mvtnorm)     # for multivariate normal distribution
library(MVN)         # for multivariate normality tests
library(ellipse)     # for generating covariance ellipses
library(ggrepel)     # for better text annotations in ggplot2
library(lavaan)      # for structural equation modeling


options(scipen = 999)
```

## Skewness 

\begin{definition}
The \emph{skewness} of a random variable $X$ is defined as the third standardized moment, given by
$$
\gamma_1 = \frac{\mu_3}{\sigma^3} = \frac{\mathbb{E}[(X - \mu)^3]}{\sigma^3},
$$
where $\mu = \mathbb{E}[X]$ is the mean and $\sigma = \sqrt{\mathbb{E}[(X - \mu)^2]}$ is the standard deviation of $X$.
\end{definition}

Skewness measures the asymmetry of a probability distribution. A distribution is said to be \emph{positively skewed} (or right-skewed) if it has a longer tail on the right side, and \emph{negatively skewed} (or left-skewed) if it has a longer tail on the left side. A perfectly symmetric distribution has a skewness of zero. To illustrate this idea, we will introduce the gamma distributions.

\begin{definition}
We say that a continuous random variable $X$ follows a \emph{Gamma distribution} with shape parameter $\alpha > 0$ and rate parameter $\beta > 0$, if its associated probability density function (PDF) is given by

\begin{equation} \label{eq:gamma_pdf}
f_X(x; \alpha, \beta) = 
\begin{cases}
\displaystyle \frac{\beta^\alpha}{\Gamma(\alpha)} \, x^{\alpha - 1} e^{-\beta x}, & x > 0, \\
0, & x \leq 0,
\end{cases}
\end{equation}
where $\Gamma(\alpha)$ denotes the Gamma function, defined as
$$
\Gamma(\alpha) = \int_0^\infty t^{\alpha - 1} e^{-t} \, dt.
$$
\end{definition}

Among other applications, the Gamma distribution is frequently used to model the number of total insurance claims over a given time period.

\begin{remark}
The skewness of a Gamma distribution is given by
$$
\gamma_1 = \frac{2}{\sqrt{\alpha}}.
$$
This means that the skewness is always positive, and it approaches zero as $\alpha$ increases. In other words, the distribution becomes more symmetric as $\alpha$ becomes larger.
\end{remark}

We will now show how to plot the PDF of the Gamma distribution for $\alpha = 2$ and $\beta = 1$, highlighting the mean and median.

```{r}
#| fig-cap: "PDF of the gamma distribution with mean and median highlighted."
#| fig-width: 8
#| fig-height: 5

# Parameters
alpha <- 2
beta <- 1

# Create sequence of x values
x_data <- seq(0, 10, length.out = 200)

# Calculate density values
gamma_distribution_tbl <- tibble(
  x_data = x_data,
  density = dgamma(x_data, shape = alpha, rate = beta)
)

# Calculate mean and median
gamma_mean <- alpha / beta
gamma_median <- qgamma(0.5, shape = alpha, rate = beta)
gamma_sd <- sqrt(alpha) / beta

# Plot
gamma_distribution_plot <- gamma_distribution_tbl %>%
  ggplot(aes(x = x_data, y = density)) +
  geom_line(
    color = c_palette["C blue"],
    linewidth = 1
  ) +
  geom_vline(
    xintercept = gamma_mean,
    color = c_palette["C red"],
    linetype = "dashed",
    linewidth = 1
  ) +
  geom_vline(
    xintercept = gamma_median,
    color = c_palette["C green"],
    linetype = "dotted",
    linewidth = 1
  ) +
  annotate(
    "text",
    x = gamma_mean,
    y = 0.5,
    label = "Mean",
    color = c_palette["C red"],
    hjust = -0.1
  ) +
  annotate(
    "text",
    x = gamma_median,
    y = 0.5,
    label = "Median",
    color = c_palette["C green"],
    hjust = 1.1
  ) +
  coord_cartesian(ylim = c(0, 0.6)) +
  labs(
    title = paste(
      "Gamma Distribution PDF (shape =", alpha, ", rate =", beta, ")"
    ),
    x = "X",
    y = "Density"
  ) +
  theme_krul()

gamma_distribution_plot
```

## Kurtosis 

\begin{definition}
The \emph{kurtosis} of a random variable $X$ is defined as the fourth standardized moment, given by
$$
\beta_2 = \frac{\mu_4}{\sigma^4} = \frac{\mathbb{E}[(X - \mu)^4]}{\sigma^4},
$$
where $\mu = \mathbb{E}[X]$ is the mean and $\sigma = \sqrt{\mathbb{E}[(X - \mu)^2]}$ is the standard deviation of $X$.
\end{definition}
Kurtosis measures the "tailedness" of a probability distribution. A distribution with high kurtosis (also known as a \emph{leptokurtic} distribution) has heavier tails and a sharper peak than a normal distribution, while a distribution with low kurtosis (also known as a \emph{platykurtic} distribution) has lighter tails and a flatter peak. The kurtosis of a normal distribution is 3, which is why we are usually interested in computing the \emph{excess kurtosis} which is given by:
$$
\gamma_2 = \beta_2 - 3.
$$
To illustrate this idea, we will introduce the Laplace distribution.
\begin{definition}
We say that a continuous random variable $X$ follows a \emph{Laplace distribution} with location parameter $\mu \in \mathbb{R}$ and scale parameter $b > 0$, if its associated probability density function (PDF) is given by
\begin{equation} \label{eq:laplace_pdf}
f_X(x; \mu, b) = \frac{1}{2b} \exp\left(-\frac{|x - \mu|}{b}\right), \quad x \in \mathbb{R}.
\end{equation}
\end{definition}

Among other applications, the Laplace distribution is often used in image processing and computer vision, particularly in modeling noise in images.

\begin{remark} The excess kurtosis of a Laplace distribution is given by
$$
\gamma_2 = 3.
$$
This means that the Laplace distribution has heavier tails than a normal distribution.
\end{remark}

We will now show how to plot the PDF of the Laplace distribution with mean zero and variance one.

```{r}
#| fig-cap: "PDF of the Laplace distribution with mean zero and variance one."
#| fig-width: 8
#| fig-height: 5

# Parameters
mu <- 0      # mean (location)
b <- 1 / sqrt(2)       # scale

# Sequence of x values covering enough range to see tails
x_data <- seq(-5, 5, length.out = 300)

# Laplace PDF function
dlaplace <- function(x, mu, b) {
  1 / (2 * b) * exp(-abs(x - mu) / b)
}

# Create data frame with PDF values
laplace_distribution_tbl <- tibble(
  x_data = x_data,
  density = dlaplace(x_data, mu, b)
)

# Plot PDF
laplace_distribution_tbl %>%
  ggplot(aes(x = x_data, y = density)) +
  geom_line(color = c_palette["C blue"], linewidth = 1) +
  labs(
    title = "Laplace Distribution PDF (mean = 0, var = 1)",
    x = "x",
    y = "Density"
  ) +
  theme_krul()
```

## QQ Plots
QQ plots, or quantile-quantile plots, are graphical tools used to compare the distribution of a dataset against a theoretical distribution, such as the normal distribution. They are particularly useful for assessing the normality of the data.

To construct a QQ plot, you should follow these steps:
\begin{enumerate}
    \item \textbf{Sort the sample data:}  
    Arrange the data in increasing order:  
    \[
    x_{(1)} \leq x_{(2)} \leq \cdots \leq x_{(n)}.
    \]
    
    \item \textbf{Calculate plotting positions (empirical quantiles):}  
    For each ordered value \( x_{(i)} \), compute the corresponding probability  
    \[
    p_i = \frac{2i - 1}{2n}, \quad i = 1, 2, \ldots, n.
    \]
    
    \item \textbf{Find theoretical quantiles of the normal distribution:}  
    Compute the theoretical quantiles from the inverse cumulative distribution function (CDF) of the normal distribution:  
    \[
    q_i = \Phi^{-1}(p_i),
    \]
    where \( \Phi^{-1} \) is the quantile function of the standard normal distribution.  
        
    \item \textbf{Plot the points:}  
    Plot the pairs \( (q_i, x_{(i)}) \) with \( q_i \) on the x-axis and \( x_{(i)} \) on the y-axis.
    
    \item \textbf{Add a reference line:}  
    Typically, we add a line through the first and third quartiles to help assess normality. \end{enumerate}

The closer the points lie to this line, the more the sample resembles the specified normal distribution.

To illustrate these ideas, we will now generate samples from the Gamma and Laplace distributions, and then we will construct their associated QQ plots and histograms. First we construct the samples:
```{r}
set.seed(123)  # for reproducibility

# Sample size
n <- 300
# Generate random samples from the Gamma distribution
gamma_sample <- rgamma(n, shape = alpha, rate = beta)

gamma_sample_tbl <- tibble(
  sample = gamma_sample
)

# Generate uniform random numbers in (0,1)
u <- runif(n)

# Apply inverse CDF of Laplace
laplace_sample <- ifelse(u < 0.5,
                         mu + b * log(2 * u),
                         mu - b * log(2 * (1 - u)))

laplace_sample_tbl <- tibble(
  sample = laplace_sample
)
```

We will now construct the corresponding QQ plots and histograms. We will start with the Gamma distribution samples:
```{r}
#| fig-cap: "QQ plot and histogram of the Gamma distribution samples."
#| fig-width: 10
#| fig-height: 5
gamma_qq_plot <- gamma_sample_tbl %>%
  ggplot(aes(sample = sample)) +
  geom_qq(
    color = c_palette["C red"],
    size = 0.3
  ) +
  geom_qq_line(
    color = c_palette["C blue"],
    linewidth = 0.5
  ) +
  labs(
    title = "QQ Plot",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_krul()

gamma_histogram <- gamma_sample_tbl %>%
  ggplot(aes(x = sample)) +
  geom_histogram(
    bins = 30,
    fill = c_palette["C blue"],
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Histogram",
    x = "Sample Values",
    y = "Density"
  ) +
  theme_krul()

gamma_plot <- gamma_qq_plot + gamma_histogram +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Gamma Distribution: QQ Plot and Histogram",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )

gamma_plot
```

We will now construct the QQ plot and histogram for the Laplace distribution samples:
```{r}
#| fig-cap: "QQ plot and histogram of the Laplace distribution samples."
#| fig-width: 10
#| fig-height: 5

laplace_qq_plot <- laplace_sample_tbl %>%
  ggplot(aes(sample = sample)) +
  geom_qq(
    color = c_palette["C red"],
    size = 0.3
  ) +
  geom_qq_line(
    color = c_palette["C blue"],
    linewidth = 0.5
  ) +
  labs(
    title = "QQ Plot",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_krul()

laplace_histogram <- laplace_sample_tbl %>%
  ggplot(aes(x = sample)) +
  geom_histogram(
    bins = 30,
    fill = c_palette["C blue"],
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Histogram",
    x = "Sample Values",
    y = "Density"
  ) +
  theme_krul()
laplace_plot <- laplace_qq_plot + laplace_histogram +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Laplace Distribution: QQ Plot and Histogram",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )
laplace_plot
```

## Sample Skewness and Kurtosis 

So far we have defined skewness and kurtosis for random variables. However, in practice, we often work with samples rather than entire populations. Therefore, we need to define sample skewness and sample kurtosis.

\begin{definition}
Let $x_1, x_2, \dots, x_n$ be a sample of size $n$. 
The \emph{sample mean} $\bar{x}$ and \emph{sample variance} $s^{2}$ are given by
\[
\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i,
\qquad
s^{2} = \frac{1}{n-1} \sum_{i=1}^n (x_i - \bar{x})^2.
\]
The \emph{sample skewness} $g_1$ is given by
\[
g_1 = \frac{n}{(n-1)(n-2)} \sum_{i=1}^n \left( \frac{x_i - \bar{x}}{s} \right)^3.
\]
The \emph{sample excess kurtosis} $g_2$ 
is given by
$$
g_2 = \frac{n(n+1)}{(n-1)(n-2)(n-3)}
\sum_{i=1}^n \left( \frac{x_i - \bar{x}}{s} \right)^4
- \frac{3(n-1)^2}{(n-2)(n-3)}.
$$
\end{definition}

As an example, we will compute the sample skewness and sample excess kurtosis of the Gamma and Laplace samples we generated earlier. For the Gamma sample we have:
```{r}
# Compute sample skewness and excess kurtosis for Gamma distribution
gamma_sample_tbl %>%
  summarise(
    sample_skewness = skewness(sample, type = 2),
    sample_excess_kurtosis = kurtosis(sample, type = 2)
  ) %>%
  glimpse()
```
For the Laplace sample we have:
```{r}
# Compute sample skewness and excess kurtosis for Laplace distribution
laplace_sample_tbl %>%
  summarise(
    sample_skewness = skewness(sample, type = 2),
    sample_excess_kurtosis = kurtosis(sample, type = 2)
  ) %>%
  glimpse()
```

## Scaling and Standardization 

In many situations, it could happen that the units we are using to measure different variables in our data are not compatible. For example, we could have a dataset that contains the height of individuals in centimeters and their weight in kilograms. In such cases, it is often useful to \emph{scale} the variables so that they are comparable. In this section, we will discuss the most common methods for scaling and standardizing random variables and samples.

### Scaling Random Variables

\begin{definition} 
An \emph{affine transformation} of a random variable $X$ is a transformation of the form 
$$
Y = aX + b
$$
where $a$ and $b$ are constants.
\end{definition}

\begin{remark}
In the context of statistics, affine transformations are often refered to as \emph{scaling} (even when they also include a translational component).
\end{remark}

Now assume that $X$ is a random variable with mean $\mu$ and standard deviation $\sigma$. Then the random variable
$$ 
Y = aX + b 
$$
will have mean $a\mu + b$ and standard deviation $|a|\sigma$. In particular, if we choose $a = 1/\sigma$ and $b = -\mu/\sigma$, then the random variable
$$
Z = \frac{X - \mu}{\sigma}
$$
will have mean $0$ and standard deviation $1$. This particular affine transformation is called a \emph{standardization}, and the resulting random variable $Z$ is called the \emph{standardized} version of $X$. $Z$ is also sometimes called the \emph{z-score} of $X$.

\begin{remark}
Notice that if $X$ is not normally distributed, then $Z$ will \emph{not} be the standard normal distribution. Standardization does not make a non-normal distribution normal.
\end{remark}

\begin{observation}
Notice that, by definition, 
$$
\gamma_{1}(Z) =\mathbb{E}[Z^{3}] = \frac{\mathbb{E}[(X - \mu)^3]}{\sigma^3} = \gamma_{1}(X)
$$
and
$$
\gamma_{2}(Z) =\mathbb{E}[Z^{4}] - 3 = \frac{\mathbb{E}[(X - \mu)^4]}{\sigma^4} -3 = \gamma_{2}(X).
$$
\end{observation}

As an example, we will consider again the Gamma distrubution that we introduced earlier. In this case we have that 
$$
\mu = \frac{\alpha}{\beta} \quad \text{and} \quad \sigma = \frac{\sqrt{\alpha}}{\beta}. 
$$
Using this information, we can generate the plot of standadized version of the Gamma distribution and compare it with the original one.
```{r}
#| fig-cap: "Comparison of the probability density function (PDF) of the original and standardized Gamma distribution."
#| fig-width: 8
#| fig-height: 5


# Calculate density values
z_gamma_distribution_tbl <- gamma_distribution_tbl %>%
  mutate(
    z_data = (x_data - gamma_mean) / gamma_sd,
    z_density = dgamma(x_data, shape = alpha, rate = beta) * gamma_sd
  ) %>%
  select(z_data, z_density)

# Calculate mean and median
z_gamma_mean <- 0
z_gamma_median <- (gamma_median - gamma_mean) / gamma_sd

# Plot
z_gamma_distribution_plot <- z_gamma_distribution_tbl %>%
  ggplot(aes(x = z_data, y = z_density)) +
  geom_line(
    color = c_palette["C blue"],
    linewidth = 1
  ) +
  geom_vline(
    xintercept = z_gamma_mean,
    color = c_palette["C red"],
    linetype = "dashed",
    linewidth = 1
  ) +
  geom_vline(
    xintercept = z_gamma_median,
    color = c_palette["C green"],
    linetype = "dotted",
    linewidth = 1
  ) +
  annotate(
    "text",
    x = z_gamma_mean,
    y = 0.55,
    label = "Mean",
    color = c_palette["C red"],
    hjust = -0.1
  ) +
  annotate(
    "text",
    x = z_gamma_median,
    y = 0.55,
    label = "Median",
    color = c_palette["C green"],
    hjust = 1.1
  ) +
  coord_cartesian(ylim = c(0, 0.6)) +
  labs(
    title = paste(
      "Standardized"
    ),
    x = "X",
    y = "Density"
  ) +
  theme_krul()

gamma_distribution_plot <- gamma_distribution_plot +
  labs(
    title = "Original"
  )

combined_gamma_plot <- gamma_distribution_plot + z_gamma_distribution_plot +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Gamma Distribution: Original and Standardized",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )

combined_gamma_plot
```


### Scaling Samples

For a given a sample, it is not always known what the underlying distribution is. In particular, we may not know what the mean and the standard deviation of the underlying population are. However, we can still try to scale our data but we should keep in mind that in this case there is more than one scaling method. The most common are:

\begin{enumerate}
    \item \emph{Standardization}: This method uses the sample mean and sample standard deviation to scale the data. The standardized sample is given by
    $$
    T = \frac{X - \bar{X}}{S},
    $$
    where $\bar{X}$ is the sample mean and $S$ is the sample standard deviation. 
    
    \item \emph{Min-Max Scaling}: This method scales the data to a fixed range, usually [0, 1]. The scaled sample is given by
    $$
    T = \frac{X - \min(X)}{\max(X) - \min(X)},
    $$
    where $\min(X)$ and $\max(X)$ are the minimum and maximum values of the sample, respectively. This method is very sensitive to outliers, as it uses the minimum and maximum values of the sample.
    
    \item \emph{Robust Scaling}: This method scales the data using the median and interquartile range (IQR). The scaled sample is given by
    $$
    T = \frac{X - \text{median}(X)}{\text{IQR}(X)},
    $$
    where $\text{IQR}(X) = Q_3 - Q_1$ is the interquartile range of the sample. This method is more robust to outliers than min-max scaling, as it uses the median and IQR instead of the minimum and maximum values.

    \item \emph{Normalization}: This method simply divides each data point by the corresponding vector norm. The normalized sample is given by
    $$
    T = \frac{X}{\|X\|},
    $$
    where $\|X\|$ is the vector norm of the sample. It is not used as often as the previous methods.
\end{enumerate}

\begin{remark} 
As we mentioned earlier, scaling does not make a non-normal distribution normal. In particular, the sample skewness and sample excess kurtosis of the scaled sample will be the same as those of the original sample.
\end{remark}

As an example, we will consider the different scaling methods applied to the Gamma sample we generated earlier.  
```{r}
#| fig-cap: "Comparison of different scaling methods applied to the Gamma distribution sample. Notice that the QQPlots are just rescaled versions of the original QQ plot, and therefore they all have the same shape."
#| fig-width: 12
#| fig-height: 14

scale_lookup_tbl <- tibble(
  code = c(
    "sample",
    "sample_standard",
    "sample_min_max",
    "sample_robust",
    "sample_normal"
  ),
  label = c(
    "Original Sample",
    "Standardized",
    "Min-Max Scaled",
    "Robust Scaled",
    "Normalized"
  )
)

scaled_gamma_sample_tbl <- gamma_sample_tbl %>%
  mutate(
    sample_standard = standard_scale(sample),
    sample_min_max = min_max_scale(sample),
    sample_robust = robust_scale(sample),
    sample_normal = normal_scale(sample)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "scaling_method",
    values_to = "sample_scaled"
  ) %>%
  convert_codes_to_factor(
    code_col = scaling_method,
    lookup_tbl = scale_lookup_tbl,
    lookup_code_col = code,
    lookup_label_col = label,
    new_factor_col_name = scaling_method_factor
  )

scaled_gamma_qq_plot <- scaled_gamma_sample_tbl %>%
  ggplot(aes(sample = sample_scaled)) +
  facet_wrap(
    ~ scaling_method_factor,
    nrow = 5,
    scales = "free"
  ) +
  geom_qq(
    color = c_palette["C red"],
    size = 0.3
  ) +
  geom_qq_line(
    color = c_palette["C blue"],
    linewidth = 0.5
  ) +
  labs(
    title = "QQ Plots",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_krul()

scaled_gamma_histogram <- scaled_gamma_sample_tbl %>%
  ggplot(aes(x = sample_scaled)) +
  facet_wrap(
    ~ scaling_method_factor,
    nrow = 5,
    scales = "free"
  ) +
  geom_histogram(
    bins = 30,
    fill = c_pal("C blue"),
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Histograms",
    x = "Sample Values",
    y = "Density"
  ) +
  theme_krul()

scaled_gamma_plot <- scaled_gamma_qq_plot + scaled_gamma_histogram +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Comparison of Scaling Methods Applied to the Gamma Sample",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )

scaled_gamma_plot
```

## Activity: Visualizing the Non-Normality of Samples

For each of the following distributions, construct a sample of size 300 and compute their associated sample Skewness and Kurtosis. After that, construct and compare their associated QQ plots and histograms.

\begin{enumerate}
    \item Exponential distribution with rate parameter $\lambda = 1$.
    \item Uniform distribution on the interval $[0, 1]$.
    \item Beta distribution with shape parameters $\alpha = 2$ and $\beta = 5$.
    \item Cauchy distribution with location parameter $x_0 = 0$ and scale parameter $\gamma = 1$.
    \item Chi-squared distribution with degrees of freedom $k = 3$.
\end{enumerate}
