# Pruebas de normalidad

Although QQ plots are a good way to visually assess the normality of a dataset, they can be subjective and imprecise. Therefore, it becomes necessary to develop more objective and precise statistical tests to determine whether the data we are analyzing follows a normal distribution. In this chapter, we will explore some commonly used statistical tests for normality and discuss some transformations that can be performed on our data to make it more normal-like. 

## Pruebas de normalidad 

There are several statistical tests that can be used to determine whether a dataset follows a normal distribution. Some of the most commonly used tests include:




\begin{enumerate}
\item \emph{Shapiro-Wilk test}: This test is based on the correlation between the data and the corresponding normal scores. It is a powerful test for small sample sizes (less than 50) and is widely used in practice. However, it can be overly sensitive to small deviations from normality in larger sample sizes. 

\item \emph{Kolmogorov-Smirnov test}: This test compares the empirical distribution function of the data with the cumulative distribution function of the normal distribution. It is a non-parametric test and can be used for any sample size. However, it is less powerful than the Shapiro-Wilk test for small sample sizes. It is also sensitive to scale and location parameters, which can affect the results of the test.

\item \emph{Anderson-Darling test}: This test is similar to the Kolmogorov-Smirnov test but gives more weight to the tails of the distribution. It is a powerful test for small sample sizes and is widely used in practice.

\item \emph{Jarque-Bera test}: This test is based on the skewness and kurtosis of the data. It is a powerful test for large sample sizes and is widely used in practice. However, it can be overly sensitive to small deviations from normality in smaller sample sizes.
\end{enumerate}

We will now apply these tests to the gamma sample data we created in the previous chapter:
```{r}
#| output: false
normality_tests_lookup_tbl <- tibble(
  code = c("shapiro", "ks", "ad", "jb"),
  label = c(
    "Shapiro-Wilk",
    "Kolmogorov-Smirnov",
    "Anderson-Darling",
    "Jarque-Bera"
  ),
)

normality_test_gamma_sample_tbl <- gamma_sample_tbl %>%
  summarise(
    shapiro = tidy_shapiro_test(data = ., value_col = sample)$p_value,
    ks = tidy_ks_test(data = ., value_col = sample)$p_value,
    ad = tidy_ad_test(data = ., value_col = sample)$p_value,
    jb = tidy_jb_test(data = ., value_col = sample)$p_value
  ) %>%
  pivot_longer(
    everything(),
    names_to = "test",
    values_to = "p_value"
  ) %>%
  convert_codes_to_factor(
    code_col = test,
    lookup_tbl = normality_tests_lookup_tbl,
    lookup_code_col = code,
    lookup_label_col = label,
    new_factor_col_name = test_factor
  ) %>%
  select(test_factor, p_value)
```
Los resultados de las pruebas de normalidad son los siguientes: 
```{r}
normality_test_gamma_sample_tbl
```
As we can see, none of these tests was particularly successful, which is not surprising given that the gamma distribution is not normal. 

## Transformaciones 
To make our data more normal-like, we can apply various transformations to any given sample. Some of the most common transformations include:
\begin{enumerate}
\item \emph{Box-Cox transformation}: This is a family of power transformations that can be applied only to data that is always positive. It is given by the formula:
$$
\tilde{X} = \begin{cases}
\frac{X^\lambda - 1}{\lambda} & \text{if } \lambda \neq 0 \\
\log(X) & \text{if } \lambda = 0 
\end{cases}
$$
where $\lambda$ is some parameter.

\item \emph{Yeo-Johnson transformation}: This transformation is similar to the Box-Cox transformation but can handle negative values. It is useful for data that is positively or negatively skewed.
It is given by the formula:
$$
\tilde{X} = \begin{cases}
\frac{((X + 1)^\lambda - 1)}{\lambda} & \text{if } X \geq 0, \lambda \neq 0 \\
\frac{-((-X + 1)^{2 - \lambda} - 1)}{2 - \lambda} & \text{if } X < 0, \lambda \neq 2 \\
\log(X + 1) & \text{if } X \geq 0, \lambda = 0 \\
\log(-X + 1) & \text{if } X < 0, \lambda = 2 
\end{cases}
$$
where $\lambda$ is some parameter. 
\end{enumerate}

We will now apply the Box-Cox transformation to the gamma sample data with parameters $\lambda = 0$, $0.25$, $0.5$, $0.75$, and $1$:

```{r}
#| fig-cap: "Comparación de distintas transformaciones de Box–Cox aplicadas a la muestra Gamma, con parámetro lambda = 0, 0.25, 0.5, 0.75 y 1."
#| fig-width: 12
#| fig-height: 14


box_cox_lookup_tbl <- tibble(
  code = c(
    "box_cox_0",
    "box_cox_025",
    "box_cox_05",
    "box_cox_075",
    "box_cox_1"
  ),
  label = c(
    "lambda = 0",
    "lambda = 0.25",
    "lambda = 0.5",
    "lambda = 0.75",
    "lambda = 1"
  ),
)

box_cox_gamma_sample_tbl <- gamma_sample_tbl %>%
  mutate(
    box_cox_0 = bcPower(sample, lambda = 0),
    box_cox_025 = bcPower(sample, lambda = 0.25),
    box_cox_05 = bcPower(sample, lambda = 0.5),
    box_cox_075 = bcPower(sample, lambda = 0.75),
    box_cox_1 = bcPower(sample, lambda = 1)
  ) %>%
  pivot_longer(
    starts_with("box_cox_"),
    names_to = "lambda",
    values_to = "transformed_sample"
  ) %>%
  convert_codes_to_factor(
    code_col = lambda,
    lookup_tbl = box_cox_lookup_tbl,
    lookup_code_col = code,
    lookup_label_col = label,
    new_factor_col_name = lambda_factor
  )


box_cox_gamma_qq_plot <- box_cox_gamma_sample_tbl %>%
  ggplot(aes(sample = transformed_sample)) +
  facet_wrap(
    ~lambda_factor,
    nrow = 5,
    scales = "free"
  ) +
  geom_qq(
    color = c_palette["C red"],
    size = 0.3
  ) +
  geom_qq_line(
    color = c_palette["C blue"],
    linewidth = 0.5
  ) +
  labs(
    title = "Gráficos QQ",
    x = "Cuantiles teóricos",
    y = "Cuantiles muestrales"
  ) +
  theme_krul()

box_cox_gamma_histogram <- box_cox_gamma_sample_tbl %>%
  ggplot(aes(x = transformed_sample)) +
  facet_wrap(
    ~lambda_factor,
    nrow = 5,
    scales = "free"
  ) +
  geom_histogram(
    bins = 30,
    fill = c_palette["C blue"],
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Histogramas",
    x = "Valores muestrales",
    y = "Densidad"
  ) +
  theme_krul()

box_cox_gamma_plot <- box_cox_gamma_qq_plot + box_cox_gamma_histogram +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Comparación de transformaciones Box–Cox con distintos valores de lambda",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )

box_cox_gamma_plot
```

## Optimización de normalidad (Box–Cox)

As we saw in the previous section, the Box-Cox transformation can be used to make our data more normal-like. However, the choice of the parameter $\lambda$ is crucial for the success of the transformation. We wish to find the optimal value of $\lambda$ that maximizes the log-likelihood of the transformed data. This can be done using the `powerTransform` function from the `car` package, which performs a Box-Cox transformation and estimates the optimal value of $\lambda$. The following figure shows the log-likelihood profile for the Box-Cox transformation applied to the gamma sample data:
```{r}
#| fig-cap: "Perfil de log-verosimilitud para la transformación de Box–Cox aplicada a la muestra Gamma. La línea discontinua indica el valor óptimo de lambda."
#| fig-width: 14
#| fig-height: 6

# Intercept-only model
fit <- lm(sample ~ 1, data = gamma_sample_tbl)

# Estimate lambda
bc <- powerTransform(fit)

# Extract log-likelihood profile
bc_prof <- boxCox(fit, lambda = seq(-2, 2, 0.1), plotit = FALSE)

# Convert to tibble for ggplot
df <- tibble(lambda = bc_prof$x, loglik = bc_prof$y)

ggplot(df, aes(x = lambda, y = loglik)) +
  geom_line(color = c_pal("C blue"), linewidth = 1) +
  geom_vline(xintercept = bc$lambda, linetype = "dashed", color = c_pal("C red")) +
  labs(
    title = "Transformación de Box–Cox",
    x = expression(lambda),
    y = "Log-verosimilitud"
  ) +
  theme_krul()
```
Finally, we can extract the optimal value of $\lambda$ and transform the data using this value:
```{r}
optimal_lambda <- bc$lambda

optimal_gamma_sample_tbl <- gamma_sample_tbl %>%
  mutate(
    transformed_sample = bcPower(sample, lambda = optimal_lambda)
  )

optimal_gamma_sample_tbl %>%
  summarise(
    shapiro = tidy_shapiro_test(data = ., value_col = transformed_sample)$p_value,
    ks = tidy_ks_test(data = ., value_col = transformed_sample)$p_value,
    ad = tidy_ad_test(data = ., value_col = transformed_sample)$p_value,
    jb = tidy_jb_test(data = ., value_col = transformed_sample)$p_value
  ) %>%
  pivot_longer(
    everything(),
    names_to = "test",
    values_to = "p_value"
  ) %>%
  convert_codes_to_factor(
    code_col = test,
    lookup_tbl = normality_tests_lookup_tbl,
    lookup_code_col = code,
    lookup_label_col = label,
    new_factor_col_name = test_factor
  ) %>%
  select(test_factor, p_value)
```
Como se observa, la transformación de Box–Cox con el valor óptimo de $\lambda$ mejora significativamente la normalidad de los datos, pues todas las pruebas arrojan valores $p$ altos. A continuación se presentan el gráfico QQ y el histograma de los datos transformados:
```{r}
#| fig-cap: "Gráfico QQ e histograma de la transformación Gamma óptima. Nótese el mejor ajuste a la normalidad en comparación con la muestra Gamma original."
#| fig-width: 10
#| fig-height: 5
optimal_gamma_qq_plot <- optimal_gamma_sample_tbl %>%
  ggplot(aes(sample = transformed_sample)) +
  geom_qq(
    color = c_palette["C red"],
    size = 0.3
  ) +
  geom_qq_line(
    color = c_palette["C blue"],
    linewidth = 0.5
  ) +
  labs(
    title = "Gráfico QQ",
    x = "Cuantiles teóricos",
    y = "Cuantiles muestrales"
  ) +
  theme_krul()

optimal_gamma_histogram <- optimal_gamma_sample_tbl %>%
  ggplot(aes(x = transformed_sample)) +
  geom_histogram(
    bins = 30,
    fill = c_palette["C blue"],
    color = "black",
    alpha = 0.7
  ) +
  labs(
    title = "Histograma",
    x = "Valores muestrales",
    y = "Densidad"
  ) +
  theme_krul()

optimal_gamma_plot <- optimal_gamma_qq_plot + optimal_gamma_histogram +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Transformación Gamma óptima",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )

optimal_gamma_plot
```
