# Análisis de componentes principales

Principal Component Analysis (PCA) is a powerful technique used to reduce the dimensionality of a dataset while preserving as much information (measured by variance) as possible. It transforms the original variables into a new set of uncorrelated variables called principal components, which are ordered by the amount of variance they explain.

## Entendiendo la varianza total

Let $\Sigma$ be the covariance matrix of a given dataset. The \emph{total variance} of the data is given by the trace of the covariance matrix: 
$$
\mbox{Total Variance} = \Tr \Sigma = \sum_{i=1}^{p} \Sigma_{ii},
$$
where $\Sigma_{ii}$ are the diagonal elements of $\Sigma$ representing the sample variance of each variable. Now, using the singular value decomposition (SVD) of $\Sigma$, we can find matrices $Q$, and $\Lambda$ such that: 
$$ 
\Sigma = Q \Lambda Q^{T}, 
$$
where $\Lambda$ is a diagonal matrix containing the singular values $\lambda_1, \ldots, \lambda_p$ of $\Sigma$. These singular values can then be used to express the total variance as:
$$
\mbox{Total Variance} = \Tr \Sigma = \Tr Q \Lambda Q^{T} = \Tr \Lambda = \lambda_1 + \cdots + \lambda_p.
$$
The columns of $Q$ are known as the \emph{principal component directions} of our data and the singular values in $\Lambda$ indicate how much variance each principal component explains. The first principal component (associated with the largest singular value) captures the most variance, the second principal component (associated with the second largest singular value) captures the second most variance, and so on. These principal components induce a new coordinate system for our data, in other words, each data point can now be represented using its \emph{principal component score}, which are related to the original variables by the relations:
$$ 
 X = Q\, (PC) + \bar{x} 
$$
and 
\begin{equation} 
PC = Q^{T} (X - \bar{x}) \label{eq:pc_scores}
\end{equation}
Here $X$ is our original coordinate vector, $PC$ is the coordinate vector corresponding to the principal component scores,  and $\bar{x}$ is the vector of means of our dataset.

To illustrate these ideas, we will consider again the dataset shown in the following figure:
```{r}
#| fig-cap: "Datos de muestra con dos variables altamente correlacionadas."
#| fig-width: 5
#| fig-height: 5
sample_data_plot
```
Recall that this dataset consists of 300 observations of two variables, $X$ and $Y$, the first one generated from a uniform distribution between $0$ and $10$, and the second one generated as $Y = X + \epsilon$, where $\epsilon$ is a normally distributed random error with mean $0$ and standard deviation $1$. The covariance matrix of this dataset is:
```{r}
Sigma <- cov(sample_data_tbl)
```
$$
\Sigma = \begin{bmatrix} 
  `r Sigma[1,1]` & `r Sigma[1,2]` \\ 
  `r Sigma[2,1]` & `r Sigma[2,2]` 
\end{bmatrix}
$$
The total variance is:
$$
\mbox{Total Variance} = \Tr \Sigma = \Sigma_{11} + \Sigma_{22} = `r Sigma[1,1] + Sigma[2,2]` 
$$
  We now look at the singular value decomposition of $\Sigma$, to obtain:
```{r}
# Compute singular vectors and values
eig <- eigen(Sigma)
vectors <- eig$vectors
values <- eig$values
``` 
$$
Q = \left[
\begin{array}{cc}
 `{r} vectors[1, 1]` & `{r} vectors[1, 2]` \\
 `{r} vectors[2, 1]` & `{r} vectors[2, 2]`
\end{array}
\right]
$$
and 
$$
\Lambda = \left[
\begin{array}{cc}
 `{r} values[1]` & 0 \\
0 & `{r} values[2]`
\end{array}
\right]
$$
The principal component directions are given by the columns of $Q$, and the amount of variance explained by each component is given by the corresponding singular value in $\Lambda$. In particular, since the vector of means is given by: 
```{r}
sample_mean_vector <- colMeans(sample_data_tbl)
```
\begin{align*}
\bar{x} & = `r sample_mean_vector[1]`  \\
\bar{y} & = `r sample_mean_vector[2]`
\end{align*}
we have that the relation between the original variables and the principal component coordinates is given by: 
\begin{align*}
X & = `r vectors[1,1]` PC_1 + `r vectors[1,2]` PC_2 + `r sample_mean_vector[1]` \\
Y & = `r vectors[2,1]` PC_1 + `r vectors[2,2]` PC_2 + `r sample_mean_vector[2]` 
\end{align*}
and 
\begin{align*} 
PC_1 & = `r vectors[1,1]` (X - `r sample_mean_vector[1]`) + `r vectors[2,1]` (Y - `r sample_mean_vector[2]`) \\ 
PC_2 & = `r vectors[1,2]` (X - `r sample_mean_vector[1]`) + `r vectors[2,2]` (Y - `r sample_mean_vector[2]`) 
\end{align*}
Using these formulas, we can find the projection of our original data into the first and second principal components. The results are shown in the following figure:

```{r}
#| fig-cap: "Proyecciones sobre las direcciones de los componentes principales. Se pierde mucha más información al proyectar sobre el segundo componente."
#| fig-width: 16
#| fig-height: 8
sample_data_matrix <- as.matrix(sample_data_tbl)

# Center the data
centered_sample_data <- scale(sample_data_matrix, center = TRUE, scale = FALSE)

# SVD
svd_result <- svd(centered_sample_data)
v1 <- svd_result$v[, 1]
v2 <- svd_result$v[, 2]

# Projection onto PC, in centered coordinates
proj_PC1 <- (centered_sample_data %*% v1) %*% t(v1)
proj_PC2 <- (centered_sample_data %*% v2) %*% t(v2)

# Shift back to original location
proj_coords1 <- proj_PC1 + attr(centered_sample_data, "scaled:center")
proj_coords2 <- proj_PC2 + attr(centered_sample_data, "scaled:center")

# Provide unique names before converting to tibble
colnames(proj_coords1) <- c("x_proj1", "y_proj1")
colnames(proj_coords2) <- c("x_proj2", "y_proj2")

proj_coords1_tbl <- as_tibble(proj_coords1)
proj_coords2_tbl <- as_tibble(proj_coords2)

# Bind with original data
compare_proj1_tbl <- bind_cols(sample_data_tbl, proj_coords1_tbl)
compare_proj2_tbl <- bind_cols(sample_data_tbl, proj_coords2_tbl)

# Generate plot corresponding to the first projection
compare_plot1 <- ggplot(compare_proj1_tbl) +
  geom_point(
    aes(x = x, y = y),
    color = "black",
    size = 0.2,
    alpha = 1
  ) +
  geom_point(
    aes(x = x_proj1, y = y_proj1),
    color = c_pal("C red"),
    size = 0.2,
    alpha = 1
  ) +
  labs(
    title = "Proyección en el primer componente principal",
    x = "X",
    y = "Y"
  ) +
  coord_fixed() +
  theme_krul()

# Generate plot corresponding to the second projection
compare_plot2 <- ggplot(compare_proj2_tbl) +
  geom_point(
    aes(x = x, y = y),
    color = "black",
    size = 0.2,
    alpha = 1
  ) +
  geom_point(
    aes(x = x_proj2, y = y_proj2),
    color = c_pal("C red"),
    size = 0.2,
    alpha = 1
  ) +
  labs(
    title = "Proyección en el segundo componente principal",
    x = "X",
    y = "Y"
  ) +
  coord_fixed() +
  theme_krul()

# Combine plots
compare_plot1 + compare_plot2 +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Proyecciones de componentes principales",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )
```

## Varianza acumulada y gráficos scree

As we mentioned in the introduction to this chapter, PCA is often used for dimensionality reduction. This means that we want to reduce the number of variables in our dataset while retaining as much information as possible. To decide how many principal components to keep, we can look at the amount of variance explained by each component. A common way to visualize this information is through the use of a \emph{scree plot}. The scree plot associated to a given dataset displays the singular values on the $y$-axis and the component number on the $x$-axis. It helps to identify the point where the explained variance starts to level off, often referred to as the "elbow". Another useful plot is the \emph{cumulative variance plot}, which shows the cumulative proportion of variance explained by the principal components. Notice that the proportion of variance explained by the $k$-th principal component is given by: 
$$ 
\frac{\lambda_k}{\sum_{i=1}^{p} \lambda_i}
$$
This plot helps to determine how many components are needed to explain a desired amount of total variance, such as $70\%$ or $90\%$. 


To illustrate this, we will use the well-known "iris" dataset, which contains measurements of four features (sepal length, sepal width, petal length, and petal width) for three species of iris flowers (setosa, versicolor, and virginica). We will perform PCA on the four features and create a scree plot to visualize the variance explained by each principal component. First, we load the data and compute the covariance matrix for each species: 
```{r}
# Load the iris dataset
data(iris)

iris_setosa_tbl <- as_tibble(iris) %>%
  filter(Species == "setosa") %>%
  select(-Species)

iris_versicolor_tbl <- as_tibble(iris) %>%
  filter(Species == "versicolor") %>%
  select(-Species)

iris_virginica_tbl <- as_tibble(iris) %>%
  filter(Species == "virginica") %>%
  select(-Species)

# Compute covariance matrix
cov_matrix_setosa <- cov(iris_setosa_tbl)
cov_matrix_versicolor <- cov(iris_versicolor_tbl)
cov_matrix_virginica <- cov(iris_virginica_tbl)

# Singular value decomposition
eig_setosa <- eigen(cov_matrix_setosa)
eig_versicolor <- eigen(cov_matrix_versicolor)
eig_virginica <- eigen(cov_matrix_virginica)
```
The covariance matrices for each species are given by: 
\begin{align*}
\Sigma_{setosa} & = \begin{bmatrix} 
  `r round(cov_matrix_setosa[1,1], 2)` & `r round(cov_matrix_setosa[1,2], 2)` & `r round(cov_matrix_setosa[1,3], 2)` & `r round(cov_matrix_setosa[1,4], 2)` \\ 
  `r round(cov_matrix_setosa[2,1], 2)` & `r round(cov_matrix_setosa[2,2], 2)` & `r round(cov_matrix_setosa[2,3], 2)` & `r round(cov_matrix_setosa[2,4], 2)` \\ 
  `r round(cov_matrix_setosa[3,1], 2)` & `r round(cov_matrix_setosa[3,2], 2)` & `r round(cov_matrix_setosa[3,3], 2)` & `r round(cov_matrix_setosa[3,4], 2)` \\ 
  `r round(cov_matrix_setosa[4,1], 2)` & `r round(cov_matrix_setosa[4,2], 2)` & `r round(cov_matrix_setosa[4,3], 2)` & `r round(cov_matrix_setosa[4,4], 2)`
\end{bmatrix} \\
\Sigma_{versicolor} & = \begin{bmatrix} 
  `r round(cov_matrix_versicolor[1,1], 2)` & `r round(cov_matrix_versicolor[1,2], 2)` & `r round(cov_matrix_versicolor[1,3], 2)` & `r round(cov_matrix_versicolor[1,4], 2)` \\ 
  `r round(cov_matrix_versicolor[2,1], 2)` & `r round(cov_matrix_versicolor[2,2], 2)` & `r round(cov_matrix_versicolor[2,3], 2)` & `r round(cov_matrix_versicolor[2,4], 2)` \\ 
  `r round(cov_matrix_versicolor[3,1], 2)` & `r round(cov_matrix_versicolor[3,2], 2)` & `r round(cov_matrix_versicolor[3,3], 2)` & `r round(cov_matrix_versicolor[3,4], 2)` \\ 
  `r round(cov_matrix_versicolor[4,1], 2)` & `r round(cov_matrix_versicolor[4,2], 2)` & `r round(cov_matrix_versicolor[4,3], 2)` & `r round(cov_matrix_versicolor[4,4], 2)`
\end{bmatrix} \\
\Sigma_{virginica} & = \begin{bmatrix} 
  `r round(cov_matrix_virginica[1,1], 2)` & `r round(cov_matrix_virginica[1,2], 2)` & `r round(cov_matrix_virginica[1,3], 2)` & `r round(cov_matrix_virginica[1,4], 2)` \\ 
  `r round(cov_matrix_virginica[2,1], 2)` & `r round(cov_matrix_virginica[2,2], 2)` & `r round(cov_matrix_virginica[2,3], 2)` & `r round(cov_matrix_virginica[2,4], 2)` \\ 
  `r round(cov_matrix_virginica[3,1], 2)` & `r round(cov_matrix_virginica[3,2], 2)` & `r round(cov_matrix_virginica[3,3], 2)` & `r round(cov_matrix_virginica[3,4], 2)` \\ 
  `r round(cov_matrix_virginica[4,1], 2)` & `r round(cov_matrix_virginica[4,2], 2)` & `r round(cov_matrix_virginica[4,3], 2)` & `r round(cov_matrix_virginica[4,4], 2)`
\end{bmatrix}
\end{align*}
The singular values for each species are given by: 
\begin{align*}
\Lambda_{setosa} & = \begin{bmatrix} 
  `r round(eig_setosa$values[1], 2)` & 0 & 0 & 0 \\ 
  0 & `r round(eig_setosa$values[2], 2)` & 0 & 0 \\ 
  0 & 0 & `r round(eig_setosa$values[3], 2)` & 0 \\ 
  0 & 0 & 0 & `r round(eig_setosa$values[4], 2)`
\end{bmatrix} \\
\Lambda_{versicolor} & = \begin{bmatrix} 
  `r round(eig_versicolor$values[1], 2)` & 0 & 0 & 0 \\ 
  0 & `r round(eig_versicolor$values[2], 2)` & 0 & 0 \\ 
  0 & 0 & `r round(eig_versicolor$values[3], 2)` & 0 \\ 
  0 & 0 & 0 & `r round(eig_versicolor$values[4], 2)`
\end{bmatrix} \\
\Lambda_{virginica} & = \begin{bmatrix} 
  `r round(eig_virginica$values[1], 2)` & 0 & 0 & 0 \\ 
  0 & `r round(eig_virginica$values[2], 2)` & 0 & 0 \\ 
  0 & 0 & `r round(eig_virginica$values[3], 2)` & 0 \\ 
  0 & 0 & 0 & `r round(eig_virginica$values[4], 2)`
\end{bmatrix}
\end{align*}
The associated principal component directions are given by the columns of the matrices
\begin{align*}
Q_{setosa} & = \left[
\begin{array}{cccc}
 `{r} round(eig_setosa$vectors[1, 1], 2)` & `{r} round(eig_setosa$vectors[1, 2], 2)` & `{r} round(eig_setosa$vectors[1, 3], 2)` & `{r} round(eig_setosa$vectors[1, 4], 2)` \\
 `{r} round(eig_setosa$vectors[2, 1], 2)` & `{r} round(eig_setosa$vectors[2, 2], 2)` & `{r} round(eig_setosa$vectors[2, 3], 2)` & `{r} round(eig_setosa$vectors[2, 4], 2)` \\
 `{r} round(eig_setosa$vectors[3, 1], 2)` & `{r} round(eig_setosa$vectors[3, 2], 2)` & `{r} round(eig_setosa$vectors[3, 3], 2)` & `{r} round(eig_setosa$vectors[3, 4], 2)` \\
 `{r} round(eig_setosa$vectors[4, 1], 2)` & `{r} round(eig_setosa$vectors[4, 2], 2)` & `{r} round(eig_setosa$vectors[4, 3], 2)` & `{r} round(eig_setosa$vectors[4, 4], 2)`
\end{array}
\right] \\
Q_{versicolor} & = \left[
\begin{array}{cccc}
 `{r} round(eig_versicolor$vectors[1, 1], 2)` & `{r} round(eig_versicolor$vectors[1, 2], 2)` & `{r} round(eig_versicolor$vectors[1, 3], 2)` & `{r} round(eig_versicolor$vectors[1, 4], 2)` \\
 `{r} round(eig_versicolor$vectors[2, 1], 2)` & `{r} round(eig_versicolor$vectors[2, 2], 2)` & `{r} round(eig_versicolor$vectors[2, 3], 2)` & `{r} round(eig_versicolor$vectors[2, 4], 2)` \\
 `{r} round(eig_versicolor$vectors[3, 1], 2)` & `{r} round(eig_versicolor$vectors[3, 2], 2)` & `{r} round(eig_versicolor$vectors[3, 3], 2)` & `{r} round(eig_versicolor$vectors[3, 4], 2)` \\
 `{r} round(eig_versicolor$vectors[4, 1], 2)` & `{r} round(eig_versicolor$vectors[4, 2], 2)` & `{r} round(eig_versicolor$vectors[4, 3], 2)` & `{r} round(eig_versicolor$vectors[4, 4], 2)`
\end{array}
\right] \\
Q_{virginica} & = \left[
\begin{array}{cccc}
 `{r} round(eig_virginica$vectors[1, 1], 2)` & `{r} round(eig_virginica$vectors[1, 2], 2)` & `{r} round(eig_virginica$vectors[1, 3], 2)` & `{r} round(eig_virginica$vectors[1, 4], 2)` \\
 `{r} round(eig_virginica$vectors[2, 1], 2)` & `{r} round(eig_virginica$vectors[2, 2], 2)` & `{r} round(eig_virginica$vectors[2, 3], 2)` & `{r} round(eig_virginica$vectors[2, 4], 2)` \\
 `{r} round(eig_virginica$vectors[3, 1], 2)` & `{r} round(eig_virginica$vectors[3, 2], 2)` & `{r} round(eig_virginica$vectors[3, 3], 2)` & `{r} round(eig_virginica$vectors[3, 4], 2)` \\
 `{r} round(eig_virginica$vectors[4, 1], 2)` & `{r} round(eig_virginica$vectors[4, 2], 2)` & `{r} round(eig_virginica$vectors[4, 3], 2)` & `{r} round(eig_virginica$vectors[4, 4], 2)`
\end{array}
\right]
\end{align*}


The scree plot and cumulative variance plot for the three species are shown in the following figure:
```{r}
#| fig-cap: "Principal Component Analysis of the Iris Dataset by Species. To decide the number of components to keep, we can use the scree plot (left) and look for the 'elbow' where the explained variance levels off. Alternatively, we can use the cumulative variance plot (right) to determine how many components are needed to explain a desired amount of total variance, such as 70% or 90%."
#| fig-width: 16
#| fig-height: 8

# Scree plot data
scree_data <- tibble(
  Component = 1:4,
  Setosa = eig_setosa$values,
  Versicolor = eig_versicolor$values,
  Virginica = eig_virginica$values
) %>%
  pivot_longer(
    cols = -Component,
    names_to = "Species",
    values_to = "SingularValue"
  )
# Scree plot
scree_plot <- scree_data %>%
  ggplot(aes(x = Component, y = SingularValue, color = Species)) +
  geom_point(size = 3) +
  geom_line() +
  c_scale_color("C red", "C blue", "C green") +
  scale_x_continuous(breaks = 1:4) +
  labs(
    title = "Gráfico scree del conjunto Iris",
    x = "Componente principal",
    y = "Valor singular"
  ) +
  theme_krul() +
  theme(legend.position = "top")

# Cumulative variance data
cumulative_variance_data <- scree_data %>%
  group_by(Species) %>%
  arrange(Component) %>%
  mutate(CumulativeVariance = cumsum(SingularValue) / sum(SingularValue))
# Cumulative variance plot
cumulative_variance_plot <- cumulative_variance_data %>%
  ggplot(aes(x = Component, y = CumulativeVariance, color = Species)) +
  geom_point(size = 3) +
  geom_line() +
  c_scale_color("C red", "C blue", "C green") +
  scale_x_continuous(breaks = 1:4) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Varianza acumulada explicada por los componentes principales",
    x = "Componente principal",
    y = "Varianza acumulada explicada"
  ) +
  theme_krul() +
  theme(legend.position = "top")

# Combine scree plot and cumulative variance plot
scree_plot + cumulative_variance_plot +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Gráfico scree y varianza acumulada para el conjunto Iris",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )
```

## Cargas (loadings) y biplots 

Recall that equation (\ref{eq:pc_scores}) gives the relation between the original variables and the principal component scores. Using this equation, we can quickly compute these scores for each observation in our dataset. The coefficients appearing in this equation are known as the \emph{loadings} of the principal components, and they indicate how much each original variable contributes to each principal component. A\emph{biplot} is a graphical representation that displays both the principal component scores of the observations and the loadings of the variables on the same plot. This allows us to visualize how the original variables contribute to the principal components and how the observations are distributed in the space defined by these components.

To illustrate this concept, we will create biplots for the first two principal components of the "iris" dataset for each species. We start by finding the loadings corresponding to each species, for which we simply take the first two columns of the matrices $Q_{setosa}$, $Q_{versicolor}$, and $Q_{virginica}$ shown above. 
\begin{align*}
\text{Loadings}_{setosa} = \left[
\begin{array}{cc}
 `{r} round(eig_setosa$vectors[1, 1], 2)` & `{r} round(eig_setosa$vectors[1, 2], 2)` \\
 `{r} round(eig_setosa$vectors[2, 1], 2)` & `{r} round(eig_setosa$vectors[2, 2], 2)` \\
 `{r} round(eig_setosa$vectors[3, 1], 2)` & `{r} round(eig_setosa$vectors[3, 2], 2)` \\
 `{r} round(eig_setosa$vectors[4, 1], 2)` & `{r} round(eig_setosa$vectors[4, 2], 2)`
\end{array}
\right]
\\ 
\text{Loadings}_{versicolor} = \left[
\begin{array}{cc}
 `{r} round(eig_versicolor$vectors[1, 1], 2)` & `{r} round(eig_versicolor$vectors[1, 2], 2)` \\
 `{r} round(eig_versicolor$vectors[2, 1], 2)` & `{r} round(eig_versicolor$vectors[2, 2], 2)` \\
 `{r} round(eig_versicolor$vectors[3, 1], 2)` & `{r} round(eig_versicolor$vectors[3, 2], 2)` \\
 `{r} round(eig_versicolor$vectors[4, 1], 2)` & `{r} round(eig_versicolor$vectors[4, 2], 2)`
\end{array}
\right]
\\
\text{Loadings}_{virginica} = \left[
\begin{array}{cc}
 `{r} round(eig_virginica$vectors[1, 1], 2)` & `{r} round(eig_virginica$vectors[1, 2], 2)` \\
 `{r} round(eig_virginica$vectors[2, 1], 2)` & `{r} round(eig_virginica$vectors[2, 2], 2)` \\
 `{r} round(eig_virginica$vectors[3, 1], 2)` & `{r} round(eig_virginica$vectors[3, 2], 2)` \\
 `{r} round(eig_virginica$vectors[4, 1], 2)` & `{r} round(eig_virginica$vectors[4, 2], 2)`
\end{array}
\right]
\end{align*}
We can then compute the principal component scores for each observation by simply multiplying the values of our original variables by these matrices. The resulting biplots for each species are shown in the following figure:
```{r}
#| fig-cap: "Cargas y puntuaciones para los dos primeros componentes principales del conjunto Iris por especie."
#| fig-width: 16
#| fig-height: 8
# Function to prepare biplot data from precomputed singular vectors
prepare_biplot_from_eigen <- function(tbl, eig) {
  data_matrix <- as.matrix(tbl)
  centered_data_matrix <- scale(data_matrix, center = TRUE, scale = FALSE)


  # Scores: project observations onto PC1 and PC2
  scores <- centered_data_matrix %*% eig$vectors[, 1:2]
  colnames(scores) <- c("PC1", "PC2")
  scores_tbl <- as_tibble(scores) %>%
    mutate(Obs = row_number())


  # Loadings: first two singular vectors
  loadings <- eig$vectors[, 1:2]
  colnames(loadings) <- c("PC1", "PC2")
  loadings <- as_tibble(loadings) %>%
    mutate(Variable = colnames(tbl))

  list(scores = scores_tbl, loadings = loadings)
}

# Prepare data for each species
setosa_data <- prepare_biplot_from_eigen(
  iris_setosa_tbl,
  eig_setosa
)

versicolor_data <- prepare_biplot_from_eigen(
  iris_versicolor_tbl,
  eig_versicolor
)

virginica_data <- prepare_biplot_from_eigen(
  iris_virginica_tbl,
  eig_virginica
)

# Function to create biplot
create_biplot <- function(scores_tbl, loadings_tbl, species_name) {
  ggplot() +
    geom_point(
      data = scores_tbl,
      aes(x = PC1, y = PC2),
      color = c_pal("C red"),
      alpha = 1,
      size = 0.3
    ) +
    geom_segment(
      data = loadings_tbl,
      aes(
        x = 0,
        y = 0,
        xend = PC1 * max(scores_tbl$PC1),
        yend = PC2 * max(scores_tbl$PC2)
      ),
      arrow = arrow(length = unit(0.3, "cm")),
      color = c_pal("C blue"),
    ) +
    geom_text(
      data = loadings_tbl,
      aes(
        x = PC1 * max(scores_tbl$PC1) * 1.1,
        y = PC2 * max(scores_tbl$PC2) * 1.1,
        label = Variable
      ),
      color = "black",
      size = 5
    ) +
    labs(
      title = paste("Biplot para", species_name),
      x = "PC1",
      y = "PC2"
    ) +
    theme_minimal()
}

# Create biplots
biplot_setosa <- create_biplot(
  setosa_data$scores,
  setosa_data$loadings,
  "Setosa"
)

biplot_versicolor <- create_biplot(
  versicolor_data$scores,
  versicolor_data$loadings,
  "Versicolor"
)

biplot_virginica <- create_biplot(
  virginica_data$scores,
  virginica_data$loadings,
  "Virginica"
)

# Combine biplots
biplot_setosa + biplot_versicolor + biplot_virginica +
  plot_layout(ncol = 3) +
  plot_annotation(
    title = "Biplots de especies de Iris (PC1 vs PC2)",
    theme = theme(
      plot.title = element_text(size = 24, face = "bold")
    )
  )
```

## Resumen del proceso de PCA

In general, when performing PCA, it is recommended to follow these steps:


\begin{enumerate}
\item \textbf{Standardize the data (if needed).} It is not always easy to decide whether to standardize the data or not. Here are some guidelines:
    \begin{itemize}
      \item If the variables are measured on different scales (e.g., height in cm and weight in kg), standardization is usually recommended.
      \item If the variables are on the same scale and have similar variances, standardization may not be necessary.
      \item If the variables have very different variances, standardization can help to give equal weight to all variables.
    \end{itemize}

  \item \textbf{Compute the covariance matrix.}
    \begin{itemize}
      \item The matrix captures how variables vary together.
      \item Its size is $p \times p$, where $p$ is the number of variables.
    \end{itemize}

  \item \textbf{Find the singular value decomposition (SVD) of the covariance matrix.}
    \begin{itemize}
      \item Eigenvalues measure the variance explained by each principal component.
      \item Eigenvectors give the directions (linear combinations of variables) of the components.
    \end{itemize}

  \item \textbf{Rank the components.}
    \begin{itemize}
      \item Order the singular values from largest to smallest.
      \item The singular vector corresponding to the largest singular value is the first principal component.
    \end{itemize}

  \item \textbf{Decide how many components to keep.}
    \begin{itemize}
      \item \textit{Kaiser criterion}: keep the singular values $> 1$ (for correlation matrix).
      \item \textit{Scree plot}: look for the “elbow” where explained variance levels off.
      \item \textit{Cumulative variance}: keep enough components to explain about $70\%-90\%$ of the total variance.
    \end{itemize}

  \item \textbf{Find the principal component scores and loadings.}
    \begin{itemize}
      \item Project the original data onto the selected principal components.
      \item Use the formula: $PC = Q^{T} (X - \bar{x})$.
    \end{itemize}

  \item \textbf{Interpret the components.}
    \begin{itemize}
      \item Examine the \textit{loadings} to see which variables contribute the most.
      \item Larger absolute values indicate stronger influence.
      \item Use biplots to visualize scores and loadings together.
      \item Apply dimension reduction before regression, clustering, or other analyses.
      \item Apply noise reduction by discarding low-variance components.
    \end{itemize}
\end{enumerate}
