# Análisis factorial: métodos prácticos

Recordemos del capítulo anterior que el modelo clásico de análisis factorial es 
$$
X = \mu + LF + \epsilon 
$$
donde:
\begin{itemize}
    \item $X$ is a random vector of $p$ observed variables (indicators).
    \item $\mu$ is a constant vector of dimension $p$.
    \item $L$ is a $p \times m$ constant matrix known as the matrix of \emph{factor loadings.}
    \item $F \sim \N(0,I_{m})$ is a vector of $m$ latent factors (unobserved variables) that are assumed to be normally distributed with mean zero and identity covariance matrix.
    \item $\epsilon \sim \N(0,\Psi)$ is a vector of unique errors (specific variances) that are also assumed to be normally distributed with mean zero and diagonal covariance matrix $\Psi = \diag(\sigma_{1}^{2}, \ldots, \sigma_{p}^{2})$.
    \item The factors $F$ and the errors $\epsilon$ are assumed to be independent, in particular $\Cov[F,\epsilon] = 0$.
\end{itemize}
Nótese que, como $\epsilon \sim \N(0,\Psi)$, este modelo queda completamente determinado por $\mu$, $L$ y $\Psi$. En este capítulo discutiremos métodos prácticos para estimar estos parámetros a partir de datos y para obtener las puntuaciones factoriales de $F$ una vez estimado el modelo.


## Evaluación de la factorizabilidad 

El primer paso es evaluar si los datos son adecuados para AF; es decir, si las variables observadas exhiben correlaciones suficientes que justifiquen un modelo factorial. Varias pruebas y medidas pueden emplearse:

\begin{enumerate}
    \item \textbf{Revisión de la matriz de correlaciones:} inspección visual en busca de patrones altos (p.ej., > 0.3) que sugieran factores comunes.
    \item \textbf{Prueba de esfericidad de Bartlett:} contrasta identidad; un resultado significativo ($p<0.05$) sugiere correlación suficiente.
    \item \textbf{Kaiser–Meyer–Olkin (KMO):} mide la proporción de varianza atribuible a factores comunes; valores cercanos a 1 indican utilidad (típicamente > 0.6 es aceptable).
\end{enumerate}

## Decidir el número de factores 

Determinar cuántos factores retener es crucial y puede afectar los resultados e interpretaciones. Algunos métodos:

\begin{enumerate}
    \item \textbf{Criterio de Kaiser:} retener valores singulares > 1.
    \item \textbf{Gráfico scree:} buscar el “codo”.
    \item \textbf{Análisis paralelo:} comparar con datos aleatorios del mismo tamaño.
    \item \textbf{Índices de ajuste:} (RMSEA, TLI) para evaluar modelos con distinto número de factores.
\end{enumerate}

## Extracción de factores

Una vez decidido el número de factores, el siguiente paso es extraerlos. Métodos disponibles:
\begin{enumerate}
    \item \textbf{Ejes principales (PAF):} se centra en la varianza compartida; estima comunalidades y extrae factores desde la matriz reducida.
    \item \textbf{Máxima verosimilitud (ML):} estima cargas maximizando la verosimilitud; permite pruebas e índices de ajuste.
\end{enumerate}

## Rotación de factores 

Tras extraer factores, suele ser útil rotarlos para lograr una estructura más simple e interpretable. La rotación puede ser ortogonal (sin correlación) u oblicua (con correlación). Métodos comunes:
\begin{enumerate}
    \item \textbf{Varimax Rotation:} An orthogonal rotation method that maximizes the variance of squared loadings within each factor, leading to a simpler structure where each variable loads highly on one factor and low on others.
    \item \textbf{Quartimax Rotation:} Another orthogonal rotation method that simplifies the rows of the loading matrix, aiming to have each variable load highly on as few factors as possible.
    \item \textbf{Equamax Rotation:} A hybrid orthogonal rotation method that combines the goals of varimax and quartimax, balancing the simplification of both rows and columns of the loading matrix.
    \item \textbf{Promax Rotation:} An oblique rotation method that allows factors to correlate. It starts with a varimax rotation and then raises the loadings to a power to achieve a simpler structure.
\end{enumerate}

## Computing Factor Scores 

Once the factor loadings have been estimated and the factors have been rotated (if applicable), the next step is to compute factor scores for each observation in the dataset. Factor scores represent the estimated values of the latent factors for each individual based on their observed variable scores. Several methods can be used to compute factor scores: 
\begin{enumerate}
    \item \textbf{Regression Method:} This method estimates factor scores by regressing the observed variables on the estimated factor loadings. It provides unbiased estimates of factor scores but may not preserve the original variance of the factors.
    \item \textbf{Bartlett Method:} This method computes factor scores by minimizing the residual variance of the observed variables. It tends to produce factor scores that are more reliable and have lower measurement error.
    \item \textbf{Anderson-Rubin Method:} This method generates factor scores that are uncorrelated and standardized. It is particularly useful when the factors are assumed to be orthogonal.
\end{enumerate}

## Assessing Model Fit 

After fitting a factor analysis model, it is important to assess how well the model fits the data. Several statistical tests and fit indices can be used to evaluate model fit: 
\begin{enumerate}
    \item \textbf{Chi-Square Test of Model Fit:} This test evaluates the null hypothesis that the model-implied covariance matrix matches the observed covariance matrix. A non-significant result ($p$-value > 0.05) suggests that the model fits the data well.
    \item \textbf{Root Mean Square Error of Approximation (RMSEA):} This index measures the discrepancy between the observed and model-implied covariance matrices per degree of freedom. Values less than 0.06 indicate a good fit.
    \item \textbf{Tucker-Lewis Index (TLI):} This index compares the fit of the specified model to a null model. Values greater than 0.95 indicate a good fit.
    \item \textbf{Standardized Root Mean Square Residual (SRMR):} This index measures the average discrepancy between observed and predicted correlations. Values less than 0.08 indicate a good fit.
\end{enumerate}

## Visualizing the Model 

Visualizing the results of a factor analysis can help in interpreting the factors and understanding the relationships among variables. Common visualization techniques include: 
\begin{enumerate}
    \item \textbf{Scree Plot:} A scree plot displays the singular values in descending order, helping to visualize the number of factors to retain.
    \item \textbf{Factor Loading Plot:} A scatter plot of factor loadings can help visualize how variables load on different factors, especially after rotation.
    \item \textbf{Biplot:} A biplot combines factor scores and loadings in a single plot, allowing for the visualization of both observations and variable relationships in the factor space.
    \item \textbf{Heatmap of Loadings:} A heatmap can be used to visualize the magnitude of factor loadings across variables and factors, highlighting patterns and clusters.
\end{enumerate}

\begin{remark}
Notice that Factor Loading Plots and Biplots only make sense when we have 2 or 3 factors, otherwise we cannot visualize them properly.
\end{remark}


## Example: Factor Analysis of the Holzinger-Swineford Dataset

We will now provide an example of factor analysis using the "HolzingerSwineford1939" dataset from the "lavaan" package. This dataset contains scores from a battery of 9 cognitive tests administered to 301 students in grades 7 and 8. The tests were designed to measure 3 latent abilities: visual/spatial, verbal, and speed. For this example we will focus only on tests x4 to x9 so that we can fit a two-factor model and visualize the results in a biplot. The 
exact description of each test is given below:

\begin{enumerate}
\item \textbf{Paragraph (x4)}: Reading comprehension. Students read short paragraphs and answer questions about the main idea, details, and simple inferences.
\item \textbf{Sentence (x5)}: Sentence completion. Choose the word or phrase that best completes a sentence so it is clear and grammatical; emphasizes vocabulary‑in‑context and syntax.
\item \textbf{Word Meaning (x6)}: Vocabulary. Select the best synonym/definition for a target word; reflects breadth of lexical knowledge.
\item \textbf{Addition (x7)}: Timed arithmetic fluency. Complete many simple addition problems quickly and accurately; captures processing speed with minimal reasoning load.
\item \textbf{Counting Dots (x8)}: Rapid visual enumeration. Quickly count small groups of dots across many items; measures visual scanning and speeded accuracy.
\item \textbf{Straight–Curved Capitals (x9)}: Perceptual discrimination speed. Scan capital letters and mark those with only straight strokes (versus any curved strokes), under strict time limits.
\end{enumerate}

To perform this analysis, we will start by loading the necessary libraries and the dataset: 
```{r}
#| output: false

# Load required libraries
library(lavaan)
library(psych)
library(scales)

# Load the Holzinger-Swineford1939 dataset
data(HolzingerSwineford1939)

# Select the relevant tests
hs_data <- as_tibble(HolzingerSwineford1939) %>%
  select("x4", "x5", "x6", "x7", "x8", "x9")
```

### Assessing Factorability

We start our factor analysis process by assessing the factorability for tests x4–x9. First we collect all the required information: 
```{r}
# Correlation matrix (pairwise complete observations)
Sigma <- cor(hs_data)

# Bartlett's test and KMO measure
n_eff <- nrow(hs_data)
bart <- cortest.bartlett(Sigma, n = n_eff)
kmo  <- KMO(Sigma)
```
Mostramos ahora el mapa de calor de correlaciones:
```{r}
#| fig-cap: "Mapa de calor de correlaciones para x4–x9. Se observan correlaciones fuertes entre las pruebas verbales (x4, x5, x6) y entre las de rapidez (x7, x8, x9), sugiriendo un modelo de dos factores."
#| fig-width: 5
#| fig-height: 5

corr_df <- Sigma %>%
  as_tibble(rownames = "Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "r")

# Preserve matrix order on axes
corr_df$Var1 <- factor(corr_df$Var1, levels = colnames(Sigma))
corr_df$Var2 <- factor(corr_df$Var2, levels = colnames(Sigma))

ggplot(corr_df, aes(x = Var2, y = Var1, fill = r)) +
  geom_tile(color = "black") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  scale_fill_gradient2(
    low = c_pal("C blue"), mid = "white", high = c_pal("C red"),
    midpoint = 0, limits = c(-1, 1)
  ) +
  coord_fixed() +
  labs(
    title = "Mapa de calor de correlaciones (x4–x9)",
    x = "", y = "", fill = "Correlación"
  ) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_blank()
  )
```
Obsérvese la fuerte correlación entre pruebas verbales (x4, x5, x6) y entre rapidez (x7, x8, x9), pero no entre grupos. Esto sugiere que un modelo de dos factores es adecuado. Reportamos ahora Bartlett y KMO:
\begin{align*}
\mbox{Bartlett's test of sphericity} & = `r bart$p.value` \\
\mbox{Kaiser-Meyer-Olkin measure} & = `r kmo$MSA`
\end{align*} 
Como se observa, todas las pruebas indican que el AF es apropiado para este conjunto.

### Decidir el número de factores 

Usaremos un gráfico scree y uno de varianza acumulada para decidir cuántos factores retener. Como usamos la matriz de correlaciones, buscamos valores singulares > 1 (Kaiser) y el "codo" en el scree.

```{r}
#| fig-cap: "Scree plot and cumulative variance for tests x4–x9. Both plots suggest that a two-factor solution is appropriate for this dataset."
#| fig-width: 14
#| fig-height: 7

# Scree analysis from the correlation matrix
eig <- eigen(Sigma)

scree_df <- tibble(
  component = seq_along(eig$values),
  singular_value = eig$values
) %>%
  mutate(
    cumulative = cumsum(singular_value) / sum(singular_value)
  )

# Scree plot
scree_plot <- scree_df %>%
  ggplot(aes(x = component, y = singular_value)) +
  geom_point(size = 3, color = c_pal("C red")) +
  geom_line(color = c_pal("C blue")) +
  geom_hline(
    yintercept = 1,
    linetype = "dashed",
    color = c_pal("C green"),
    linewidth = 1.5
  ) +
  scale_x_continuous(breaks = scree_df$component) +
  labs(
    title = "Gráfico scree",
    x = "Componente principal",
    y = "Valor singular"
  ) +
  theme_krul()

# Cumulative variance plot
cum_plot <- ggplot(scree_df, aes(x = component, y = cumulative)) +
  geom_point(size = 3, color = c_pal("C red")) +
  geom_line(color = c_pal("C blue")) +
  scale_x_continuous(breaks = scree_df$component) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  labs(
    title = "Varianza acumulada",
    x = "Componente principal",
    y = "Varianza acumulada explicada"
  ) +
  theme_krul()

# Combine scree plot and cumulative variance plot
scree_plot + cum_plot + plot_layout(ncol = 2) +
  plot_annotation(
    title = paste(
      "Gráfico scree y varianza acumulada para",
      "el conjunto Holzinger-Swineford"
    ),
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )
```
Notice that both the scree plot and the cumulative variance plot suggest that a two-factor solution is appropriate for analysing the six tests included in this dataset.

### Extracting, Rotating, and Scoring Factors

Even though extracting, rotating and scoring factors are theoretically separate steps, in practice they are often performed together using specialized software or statistical packages. Here we will use the `psych::fa` function to perform all three steps at once. We will fit a 2-factor model using maximum likelihood (ML) estimation, compute factor scores using the regression method, and perform a varimax rotation to simplify the factor structure. 

```{r} 
# Fit 2-factor FA using ML, regression scores and varimax rotation
fa_rot <- fa(
  r = hs_data,
  nfactors = 2,
  rotate = "varimax",
  fm = "ml",
  scores = "regression",
  use = "pairwise"
)

# Extract factor loadings into a tibble
loadings <- fa_rot$loadings[, 1:2] %>%
  unclass() %>%
  as_tibble(rownames = "Variable") %>%
  rename(F1 = 2, F2 = 3)
# Extract factor scores into a tibble
scores <- fa_rot$scores %>%
  as_tibble() %>%
  setNames(c("F1", "F2")) %>%
  mutate(Obs = row_number())
```

### Assessing Model Fit 

We now evaluate how well the 2‑factor ML solution reproduces the observed correlations. For this, we perform the tests $\chi$‑squared, RMSEA, CFI, TLI, and RMSR. 
```{r}
#| label: model-fit-x4-x9
#| echo: true
#| message: true
#| warning: false

# Extract standard fit indices from psych::fa object (ML)
chisq <- fa_rot$STATISTIC
df    <- fa_rot$dof
pval  <- fa_rot$PVAL
rmsea <- as.numeric(fa_rot$RMSEA[1])
tli   <- fa_rot$TLI
bic   <- fa_rot$BIC

L   <- as.matrix(fa_rot$loadings)
Phi <- if (!is.null(fa_rot$Phi)) fa_rot$Phi else diag(ncol(L))
uni <- if (!is.null(fa_rot$uniquenesses)) {
  fa_rot$uniquenesses
} else {
  1 - rowSums((L %*% Phi) * L)
}
Rhat <- L %*% Phi %*% t(L) + diag(as.numeric(uni))

# Helper function to extract lower-triangle entries (off-diagonal)
lt_offdiag <- function(M) M[lower.tri(M, diag = FALSE)]

# Classic RMSR
res <- Sigma - Rhat
rmsr <- sqrt(mean(lt_offdiag(res)^2, na.rm = TRUE))

# Standardized RMSR (SRMR)
is_cov <- any(abs(diag(Sigma) - 1) > 1e-8)
S_obs_cor  <- if (is_cov) stats::cov2cor(Sigma) else Sigma
S_hat_cor  <- if (is_cov) stats::cov2cor(Rhat)  else Rhat

res_cor <- S_obs_cor - S_hat_cor

# SRMR = sqrt( sum_{i<j}(res_ij^2) / sum_{i<j}(obs_ij^2) )
num <- sum(lt_offdiag(res_cor)^2, na.rm = TRUE)
den <- sum(lt_offdiag(S_obs_cor)^2, na.rm = TRUE)
srmr <- sqrt(num / den)
```
The results of these tests are:
\begin{align*} 
\mbox{$\chi^{2}$ Test} & = `r pval` \\ 
\mbox{RMSEA} & = `r rmsea` \\ 
\mbox{TLI} & = `r tli` \\ 
\mbox{RMSR} & = `r srmr` 
\end{align*}
All tests indicate that the 2‑factor model fits the data well.

### Visualizing the Model

Finally, we will visualize the model using a biplot. We will also rotate the factors manually to illustrate what could happen if we did not choose an appropriate rotation method. 

```{r}
#| fig-cap: "Biplots of the 2-factor solution for tests x4–x9. Left: Varimax rotation. Right: Manual 45° rotation. Notice how the varimax rotation leads to a simpler and more interpretable structure, with verbal tests (x4, x5, x6) loading highly on Factor 1 and speed tests (x7, x8, x9) loading highly on Factor 2."
#| fig-width: 14 
#| fig-height: 7

# Scaling factor for arrows (keep consistent across plots)
arrow_scale <- 3

# Base (unrotated) biplot
p_base <- ggplot() +
  geom_point(
    data = scores,
    aes(x = F1, y = F2),
    alpha = 1,
    color = c_pal("C red"),
    size = 0.5
  ) +
  geom_segment(
    data = loadings,
    aes(x = 0, y = 0, xend = F1 * arrow_scale, yend = F2 * arrow_scale),
    arrow = arrow(length = unit(0.2, "cm")),
    color = c_pal("C blue"),
    linewidth = 1.2
  ) +
  geom_text(
    data = loadings,
    aes(x = F1 * arrow_scale, y = F2 * arrow_scale, label = Variable),
    color = "black",
    hjust = -0.5
  ) +
  labs(
    title = "Varimax Rotation",
    x = "Factor 1",
    y = "Factor 2"
  ) +
  theme_krul()

# 45° rotation of scores and loadings
theta  <- pi / 4
cos_t  <- cos(theta)
sin_t  <- sin(theta)

scores_rot <- scores %>%
  mutate(
    F1_rot = F1 * cos_t - F2 * sin_t,
    F2_rot = F1 * sin_t + F2 * cos_t
  )

loadings_rot <- loadings %>%
  mutate(
    F1_rot = F1 * cos_t - F2 * sin_t,
    F2_rot = F1 * sin_t + F2 * cos_t
  )

# Rotated biplot
p_rot <- ggplot() +
  geom_point(
    data = scores_rot,
    aes(x = F1_rot, y = F2_rot),
    alpha = 1,
    color = c_pal("C red"),
    size = 0.5
  ) +
  geom_segment(
    data = loadings_rot,
    aes(x = 0, y = 0, xend = F1_rot * arrow_scale, yend = F2_rot * arrow_scale),
    arrow = arrow(length = unit(0.2, "cm")),
    color = c_pal("C blue"),
    linewidth = 1.2
  ) +
  geom_text(
    data = loadings_rot,
    aes(x = F1_rot * arrow_scale, y = F2_rot * arrow_scale, label = Variable),
    color = "black",
    hjust = -0.5
  ) +
  labs(
    title = "Manual Rotation",
    x = "Rotated Factor 1",
    y = "Rotated Factor 2"
  ) +
  theme_krul()

# Optional: share axis limits for fair comparison
xr <- range(
  c(
    scores$F1,
    scores_rot$F1_rot,
    loadings$F1 * arrow_scale,
    loadings_rot$F1_rot * arrow_scale
  ),
  na.rm = TRUE
)

yr <- range(
  c(
    scores$F2,
    scores_rot$F2_rot,
    loadings$F2 * arrow_scale,
    loadings_rot$F2_rot * arrow_scale
  ),
  na.rm = TRUE
)
p_base <- p_base + coord_equal(xlim = xr, ylim = yr)
p_rot  <- p_rot  + coord_equal(xlim = xr, ylim = yr)

# Side-by-side using patchwork
p_base + p_rot + plot_layout(ncol = 2) + 
  plot_annotation(
    title = paste(
      "Biplots of the factor analysis model",
      "using Varimax and Manual Rotations"
    ),
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )
```
