--- 
title: |
  Practice with continous variables in R,

  Histograms, Box Plots, and Density Plots

format:
  pdf:
    pdf-engine: pdflatex
    fontsize: 11pt
    number-sections: true

output-dir: reports/

highlight-style: tango 

execute:
  echo: true  # Set to false if you don't want code displayed
---

# Continuous Variables in R 

In this notes, we will practice with continuous variables in R, focusing on histograms, box plots, and density plots. We will start by loading the necessary libraries and the "diamond" dataset.
```{r}
#| output: false
library("tidyverse")
library("here")
library("patchwork")
library("krulRutils")

options(scipen = 999) # Disable scientific notation
data(diamonds)
```

## Box Plots
Box plots are a great way to visualize the distribution of continuous variables. The following figure shows the main components of a box plot:
```{r}
#| fig-width: 6
#| fig-height: 12
#| fig-cap: "\\small Description of the main components of a boxplot"
boxplot_explanation_plot <- ggplot_box_legend()
ggsave(
  filename = here("images", "boxplot_explanation.png"),
  plot = boxplot_explanation_plot,
  width = 6,
  height = 12,
  dpi = 300
)
boxplot_explanation_plot
```
