--- 
title: |
  Practice with continous variables in R,

  Histograms, Box Plots, and Density Plots

format:
  pdf:
    pdf-engine: pdflatex
    fontsize: 11pt
    number-sections: true

output-dir: reports/

highlight-style: tango 

execute:
  echo: true  # Set to false if you don't want code displayed
---

# Continuous Variables in R 

In this notes, we will practice with continuous variables in R, focusing on histograms, box plots, and density plots. We will start by loading the necessary libraries and the "diamonds" dataset.
```{r}
#| output: false
library("tidyverse")
library("here")
library("patchwork")
library("krulRutils")

options(scipen = 999) # Disable scientific notation
data(diamonds)
```
Before starting plotting our data, we first take a little glimpse at the data contained in this dataset:
```{r}
diamonds %>%
  glimpse()
```
We also look at the most important summary statistics:
```{r}
diamonds %>%
  summarise_numeric_tidy()
```

## Box Plots
Box plots are a great way to visualize the distribution of continuous variables. The following figure shows the main components of a box plot:
```{r}
#| fig-width: 6
#| fig-height: 6
#| fig-cap: "\\small Description of the main components of a boxplot"
boxplot_explanation_plot <- ggplot_box_legend()
ggsave(
  filename = here("images", "boxplot_explanation.png"),
  plot = boxplot_explanation_plot,
  width = 6,
  height = 12,
  dpi = 300
)
boxplot_explanation_plot
```
Now we will construct the boxplot associated to the "price" variable of this dataset:
```{r}
#| fig-width: 9
#| fig-height: 2
#| fig-cap: "\\small Boxplot showing the general distribution of diamond prices"
diamonds_boxplot <- diamonds %>%
  ggplot(aes(x = price, y = 1)) +
  geom_boxplot(
    color = "black",
    fill = c_pal("C cyan"),
    outlier.shape = NA
  ) +
  stat_boxplot(geom = "errorbar", width = 0.3, linewidth = 0.5) +
  geom_jitter(height = 0.3, alpha = 0.1, size = 0.1, color = c_pal("C red")) +
  coord_cartesian(xlim = c(0, 19000)) +
  labs(
    title = "Diamonds price boxplot",
    x = "Price (USD)",
    y = ""
  ) +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  theme_krul()

diamonds_boxplot
```
Now we will do a more detailed analysis by comparing the boxplots corresponding to the different categories associated to the variable "cut".
```{r}
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "\\small Boxplot showing the distribution of diamond prices by 'cut'"
diamonds_boxplot_by_cut <- diamonds %>%
  ggplot(aes(x = price, y = 1, fill = cut)) +
  geom_boxplot(
    color = "black",
    outlier.shape = NA
  ) +
  stat_boxplot(geom = "errorbar", width = 0.3, linewidth = 0.5) +
  geom_jitter(
    aes(color = cut),
    width = 0.375,
    alpha = 0.1,
    size = 0.1
  ) +
  facet_wrap(~cut, ncol = 1) +
  c_scale_fill("C spring green", "C cyan", "C azure", "C blue", "C violet") +
  c_scale_color("C rose", "C red", "C orange", "C yellow", "C chartreuse") +
  coord_cartesian(xlim = c(0, 19000)) +
  labs(
    title = "Diamonds price by cut boxplot",
    x = "Price (USD)",
    y = "",
  ) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "none"
  ) +
  theme_krul()

diamonds_boxplot_by_cut
```

## Histograms 
We will now consider the histograms associated with this data. As before we start by looking at the histogram associated to the variable "price":
```{r}
#| fig-width: 8
#| fig-height: 4.5
#| fig-cap: "\\small Histogram corresponding to the variable price"
diamonds_histogram <- diamonds %>%
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 500, color = "black", fill = c_pal("C cyan")) +
  labs(
    title = "Diamonds price histogram",
    x = "Price (USD)",
    y = "Count"
  ) +
  coord_cartesian(xlim = c(0, 19000)) +
  theme_krul()

diamonds_histogram
```
We no proceed to a more detailed analysis by breaking our plot according to the different categories associated to the variable "cut" 
```{r}
#| fig-width: 9
#| fig-height: 16
#| fig-cap: "\\small Histogram corresponding to the variable price divided according to the variable 'cut'"
diamonds_histogram_by_cut <- diamonds %>%
  ggplot(aes(x = price, fill = cut)) +
  geom_histogram(binwidth = 500, color = "black") +
  facet_wrap(~cut, ncol = 1, scales = "free_y") +
  c_scale_fill("C spring green", "C cyan", "C azure", "C blue", "C violet") +
  labs(
    title = "Histogram of the price variable by cut",
    x = "Price (USD)",
    y = "Count"
  ) +
  coord_cartesian(xlim = c(0, 19000)) +
  theme_krul() +
  theme(legend.position = "none")

diamonds_histogram_by_cut
```

## Density plots
Finally, we will analyze the density plots associated to the "price" variable. Density plots provide similar information to histograms, but instead of showing the actual counts, it tries to estimate the shape of the associated probability density funtion.
```{r}
#| fig-width: 8
#| fig-height: 4.5
#| fig-cap: "\\small Density plot corresponding to the variable price"
diamonds_density_plot <- diamonds %>%
  ggplot(aes(x = price)) +
  geom_density(color = "black", fill = c_pal("C cyan")) +
  labs(
    title = "Diamonds density plot",
    x = "Price (USD)",
    y = "Density"
  ) +
  coord_cartesian(xlim = c(0, 19000)) +
  theme_krul()

diamonds_density_plot
```
As before, we will now perform a more detailed analysis taking the "cut" variable into account
```{r}
#| fig-width: 9
#| fig-height: 16
#| fig-cap: "\\small Density plot corresponding to the variable price divided according to the variable 'cut'"
diamonds_density_plot_by_cut <- diamonds %>%
  ggplot(aes(x = price, fill = cut)) +
  geom_density(color = "black") +
  coord_cartesian(xlim = c(0, 19000)) +
  facet_wrap(~cut, ncol = 1, scale = "free_y") +
  c_scale_fill("C spring green", "C cyan", "C azure", "C blue", "C violet") +
  labs(
    title = "Diamonds density plot by cut",
    x = "Price (USD)",
    y = "Density"
  ) +
  theme_krul() +
  theme(legend.position = "none")

diamonds_density_plot_by_cut
```
Density plots associated to different categories are frequently plot together. In out case, the 5 categories make it hard to distinguish the plots when they are all shown together, but we can filter the categories to illustrate this option.
```{r}
#| fig-width: 8
#| fig-height: 4.5
#| fig-cap: "\\small Density plot corresponding to the variable price divided according to the filtered variable 'cut'"
diamonds_filtered_density_plot <- diamonds %>%
  filter(cut %in% c("Fair", "Ideal")) %>%
  ggplot(aes(x = price, fill = cut)) +
  geom_density(color = "black", alpha = 0.7) +
  coord_cartesian(xlim = c(0, 19000)) +
  c_scale_fill("C spring green", "C violet") +
  labs(
    title = "Diamonds density plot by filtered cut",
    x = "Price (USD)",
    y = "Density"
  ) +
  theme_krul() +
  theme(legend.position = "none")

diamonds_filtered_density_plot
```

# Using patchwork to put plots together

Finally in this section we will show how we can use patchwork to put all of this information together. Since we have already covered how to find all of these plots, we will just show the corresponding code.
```{r}
#| fig-width: 16
#| fig-height: 9
#| fig-cap: "\\small Plot showing the boxplots, histograms and density plots corresponding to the price of diamonds by cut"

diamonds_density_plot <- diamonds %>%
  ggplot(aes(x = price, fill = cut)) +
  facet_wrap(~cut, nrow = 1) +
  geom_density(
    alpha = 1,
    color = "black"
  ) +
  c_scale_fill("C spring green", "C cyan", "C azure", "C blue", "C violet") +
  coord_cartesian(xlim = c(0, 19000)) +
  labs(
    title = "Density Plots",
    x = "Price (USD)",
    y = "Density",
    fill = "Cut: "
  ) +
  theme_krul() +
  theme(
    strip.background = element_blank(),
    strip.text = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 10),
    legend.background = element_rect(fill = "gray95", color = "black"),
  )

diamonds_histogram <- diamonds %>%
  ggplot(aes(x = price, fill = cut)) +
  facet_wrap(~cut, nrow = 1) +
  geom_histogram(binwidth = 500, color = "black") +
  coord_cartesian(xlim = c(0, 19000)) +
  c_scale_fill("C spring green", "C cyan", "C azure", "C blue", "C violet") +
  labs(
    title = "Histograms",
    x = "Price (USD)",
    y = "Count",
    fill = "Cut"
  ) +
  theme_krul() +
  theme(
    strip.background = element_blank(),
    strip.text = element_blank(),
    legend.position = "none",
  )

diamonds_boxplot <- diamonds %>%
  ggplot(aes(x = price, y = 0, fill = cut, color = cut)) +
  facet_wrap(~cut, nrow = 1) +
  geom_boxplot(
    color = "black",
    outlier.shape = NA
  ) +
  stat_boxplot(
    geom = "errorbar",
    width = 0.3,
    linewidth = 0.5,
    color = "black"
  ) +
  geom_jitter(
    height = 0.375,
    alpha = 0.1,
    size = 0.1
  ) +
  c_scale_fill("C spring green", "C cyan", "C azure", "C blue", "C violet") +
  c_scale_color("C rose", "C red", "C orange", "C yellow", "C chartreuse") +
  coord_cartesian(xlim = c(0, 19000)) +
  labs(
    title = "Boxplots",
    x = "Price (USD)",
    y = "",
    fill = ""
  ) +
  theme_krul() +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    strip.background = element_blank(),
    strip.text = element_blank(),
    legend.position = "none"
  )

# Now stack vertically (rows) using patchwork’s `/`
diamonds_combined_plot <- diamonds_boxplot /
  diamonds_histogram /
  diamonds_density_plot

diamonds_combined_figure <- diamonds_combined_plot +
  plot_layout(heights = c(1, 3, 3)) +
  plot_annotation(
    title = "Distribution of Diamond Prices by Cut",
    theme = theme(
      plot.title = element_text(
        size = 24,
        face = "bold"
      )
    )
  )

ggsave(
  filename = here("images", "diamonds_combined_figure.png"),
  plot = diamonds_combined_figure,
  width = 16,
  height = 9,
  dpi = 300
)

diamonds_combined_figure
```
